---
title: "Initial QC & Data Cleaning"
---

## Libraries

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics')
library(data.table)
library(MSstats)
library(qs2)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(pheatmap)
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run1"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in meta and spec raw data

```{r}
meta <- fread(file.path(raw_dir, "LCM_2025_metadata.csv"))
spec_raw <- fread(file.path(raw_dir, "20260215_204256_hela20_Report.tsv"))
meta
```

## Clean metadata

Remove hidden spaces etc.

```{r}
meta_cleaned <- meta %>%
  mutate(
    # 1. Trim hidden spaces and force everything to lowercase first
    Condition = tolower(str_trim(Condition)),
    
    # 2. Map all variations directly to your snake_case targets
    Condition = case_when(
      Condition %in% c("close", "plaque near") ~ "plaque_near",
      Condition == "plaque far"               ~ "plaque_far",
      Condition == "control"                  ~ "control",
      TRUE ~ Condition # Keeps anything else exactly as is, so you can spot typos
    ),
    
    # 3. Handle the Mouse column (use backticks if it still has the trailing space from the raw CSV)
    BioReplicate = str_trim(`Mouse`) 
  ) %>%
  as.data.table()

meta_cleaned
```

### Check number of columns

```{r}
all_cols <- colnames(spec_raw)
length(all_cols)
quant_cols <- grep("PG\\.Quantity", all_cols, value = TRUE)
length(quant_cols)

run_cols <- grep("hela", quant_cols, value = TRUE, invert = TRUE)
length(run_cols)

run_mapping <- data.table(Run = run_cols)
run_mapping
length(run_mapping$Run)
```

## Extract chip location and batch

Importantly, use the machine truth section of the Run, as the typed labelling is incorrect for one sample.

BDAO5.1D in well C4 batch 5 wasn't mapped -\> misnamed, but luckily machine truth is still there in capitals at the end.

```{r}
grep("c4", colnames(spec_raw), ignore.case = TRUE, value = TRUE)
```

## Proof of mislabelling

If you look at run\[160\], the file name is **`mt_c2_S3-C4_1_5735`**.

The first part of the name (`mt_c2`) is what was manually typed into the sequence table.

The middle part of the name (`-C4_1`) is the physical autosampler well that the mass spec actually drew the sample from.

Look at the injection counter at the very end of the files. Run `[159]` was injected at time **`5733`** (from well C2). The very next sample, run `[160]`, was injected at time **`5735`** (from well C4).

```{r}
grep("S3.*[Cc][24]", colnames(spec_raw), value = TRUE)
```

## Mapping using the machine truth

```{r}
run_mapping <- run_mapping %>%
  mutate(
    # 1. MACHINE TRUTH WELL: Extract the uppercase well between the hyphen and underscore
    # e.g., matches "-C4_" and pulls out "C4"
    Chip_location = str_match(Run, "-([A-Z][0-9]+)_")[, 2],
    
    # 2. BATCH EXTRACTION: Keep this the same (pulls the number/S3 after the first well)
    Batch_raw = str_match(tolower(Run), "mt_[a-z0-9]+_([0-9]+|s3)")[, 2],
    
    # 3. FIX BATCH: Swap "s3" for "5" and convert to numeric
    Batch_raw_fixed = str_replace(Batch_raw, "s3", "5"),
    Batch = as.numeric(Batch_raw_fixed)
  ) %>%
  # Drop the temporary columns
  select(-Batch_raw, -Batch_raw_fixed) %>% 
  as.data.table()

# Quick sanity check for the problem file:
print(run_mapping[grep("5735", Run)])
```

## Merge mapped cleaned metadata to the Runs

```{r}
meta_mapped <- merge(
  meta_cleaned, 
  run_mapping, 
  by = c("Chip_location", "Batch"), 
  all.x = TRUE
)

missing_mapping <- meta_mapped[is.na(Run)]
if(nrow(missing_mapping) > 0) {
  cat("Warning: The following samples are missing from the Spectronaut columns:\n")
  print(missing_mapping[, .(BioReplicate, Chip_location, Batch)])
} else {
  cat("All samples successfully mapped!\n")
}
meta_mapped
```

## Check that all samples were correctly merged

```{r}
missing_mapping <- meta_mapped[is.na(Run)]
if(nrow(missing_mapping) > 0) {
  cat("Warning: The following samples are missing from the Spectronaut columns:\n")
  print(missing_mapping[, .(BioReplicate, Chip_location, Batch)])
} else {
  cat("All metadata samples successfully mapped!\n")
}
```

## Check for blank samples

These are removed from analysis - but could be worth referring to ?

```{r}
extra_runs <- run_mapping[!Run %in% meta_mapped$Run]
cat("These are the", nrow(extra_runs), "mass spec runs that were dropped because they weren't in the metadata:\n")
print(extra_runs)
```

## Calculate protein ID counts

From wide format.

Stores raw matrix of count data as `quant_data`

Stores [**metadata**]{.underline} as `qc_data`, after including number of counts and run name.

```{r}
#Calculates ID counts from wide format

# Extract just the valid run columns that successfully mapped to a mouse
valid_runs <- meta_mapped[!is.na(Run)]$Run

# Subset the raw data to only these quantitative columns
quant_data <- spec_raw[, ..valid_runs]

# A protein is "identified" if it has a non-NA value greater than 0
protein_counts <- colSums(!is.na(quant_data) & quant_data > 0, na.rm = TRUE)

# Create the summary table
qc_counts <- data.table(
  Run = names(protein_counts),
  Protein_Count = protein_counts
)

# Merge the biological metadata with our technical counts
qc_data <- merge(qc_counts, meta_mapped, by = "Run", all.x = TRUE)

# Print a quick summary to the console
summary(qc_data$Protein_Count)

```

## Run PCA

Data already wide. - First need to log-transform it - handle the missing values - transpose it so runs are rows.

```{r}
cat("Preparing data for PCA...\n")

# Convert the subsetted data to a numeric matrix
pca_matrix <- as.matrix(quant_data)

# Log2 transform the intensities (adding +1 to avoid log2(0))
pca_matrix_log <- log2(pca_matrix + 1)

# Transpose it so Rows = Runs, Columns = Proteins (prcomp expects this)
pca_matrix_log_t <- t(pca_matrix_log)

# 1. IMPUTE FIRST: Fill missing values (NAs or 0s) with a tiny noise floor
noise_floor <- min(pca_matrix_log_t[pca_matrix_log_t > 0], na.rm = TRUE) / 2
pca_matrix_log_t[is.na(pca_matrix_log_t) | pca_matrix_log_t == 0] <- noise_floor

# 2. FILTER SECOND: Now remove any proteins that have zero variance 
# (e.g., proteins that were all NAs and are now just a flat line of noise_floor)
valid_cols <- apply(pca_matrix_log_t, 2, var) > 0
pca_matrix_log_t <- pca_matrix_log_t[, valid_cols]

cat("Running prcomp...\n")
pca_res <- prcomp(pca_matrix_log_t, center = TRUE, scale. = TRUE)

# Extract coordinates and merge with metadata
pca_df <- as.data.frame(pca_res$x)
pca_df$Run <- rownames(pca_df)
pca_df <- merge(pca_df, meta_mapped, by = "Run", all.x = TRUE)

# Calculate % variance explained by PC1 and PC2
variance_explained <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

```

### Generate PDF report of QC metrics

-   Bar chart of total proteins identified by Run

-   Bar chart to compare protein quality by condition and cut quality

-   PCAs coloured by:

    -   Condition

    -   Cut

    -   Batch

```{r, fig.width=10, fig.height=7}

# ---------------------------------------------------------
# Plot 1: Protein Count per Run
# ---------------------------------------------------------
p1 <- ggplot(qc_data, aes(x = reorder(Run, Protein_Count), y = Protein_Count, fill = Condition)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  coord_flip() + 
  scale_fill_manual(values = c("control" = "#1b9e77", "plaque_near" = "#d95f02", "plaque_far" = "#7570b3")) +
  labs(title = "Total Proteins Identified per Run",
       subtitle = "Check for sharp drop-offs indicating failed injections",
       x = "Mass Spec Run", y = "Protein Count") +
  theme(axis.text.y = element_text(size = 4)) 

# ---------------------------------------------------------
# Plot 2: Yield by Condition & Cutting Quality
# ---------------------------------------------------------
p2 <- ggplot(qc_data, aes(x = Condition, y = Protein_Count, fill = Cutting_quality)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(width = 0.75)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75), 
             alpha = 0.6, size = 1.5) +
  theme_minimal() +
  labs(title = "Protein Yield by Condition and Cut Quality",
       x = "Biological Condition", y = "Protein Count")

# ---------------------------------------------------------
# Plot 3: PCA (Color by Condition)
# ---------------------------------------------------------
p3 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("control" = "#1b9e77", "plaque_near" = "#d95f02", "plaque_far" = "#7570b3")) +
  labs(title = "PCA of Log2 Protein Intensities",
       subtitle = "Checking for Biological Separation",
       x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)")) +
  theme(legend.position = "right")

# ---------------------------------------------------------
# Plot 4: PCA (Color by Cutting Quality)  <--- NEW PLOT
# ---------------------------------------------------------
p4 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Cutting_quality)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  # Using distinct colors so it's easy to spot the "Bad" ones
  scale_color_manual(values = c("Fine" = "gray70", "Bad" = "red")) +
  labs(title = "PCA Colored by Cut Quality",
       subtitle = "If 'Bad' cuts isolate on one side of the plot, they should be dropped",
       x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)"),
       color = "Cut Quality") +
  theme(legend.position = "right")

# ---------------------------------------------------------
# Plot 5: PCA (Color by Batch)
# ---------------------------------------------------------
p5 <- ggplot(pca_df, aes(x = PC1, y = PC2, color = as.factor(Batch))) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA Colored by Mass Spec Batch",
       subtitle = "If colors cluster perfectly, we have a Batch Effect",
       x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)"),
       color = "Batch")

# ---------------------------------------------------------
# 2. Save to PDF (The background export)
# ---------------------------------------------------------
pdf_path <- file.path(results_dir, "Initial_QC_Report.pdf")
cat("Generating PDF report at:", pdf_path, "\n")

pdf(pdf_path, width = 10, height = 7)
print(p1)
print(p2)
print(p3)
print(p4)
print(p5)
dev.off()

# ---------------------------------------------------------
# 3. Render Inline (For the Quarto viewer)
# ---------------------------------------------------------
cat("Displaying plots inline for review...\n")
print(p1)
print(p2)
print(p3)
print(p4)
print(p5)
```

## MAD outlier detection

Code here identifies samples that have 3 MAD below median protein counts.

Stores samples to drop in `failed_runs` variable.

```{r}
cat("Calculating MAD thresholds for Protein Counts...\n")

# Calculate Median and MAD
med_count <- median(qc_data$Protein_Count, na.rm = TRUE)
# R's mad() function automatically applies the 1.4826 scaling factor to approximate SD
mad_count <- mad(qc_data$Protein_Count, na.rm = TRUE) 

# Set the cutoff at 3 MADs below the median
mad_cutoff <- med_count - (3 * mad_count)

cat("Median Protein Count:", round(med_count), "\n")
cat("MAD:", round(mad_count), "\n")
cat("Dropping any run below:", round(mad_cutoff), "proteins\n")

# Identify the exact runs that failed
failed_runs <- qc_data %>%
  filter(Protein_Count < mad_cutoff) %>%
  pull(Run)

if(length(failed_runs) > 0) {
  cat("\nWARNING: Found", length(failed_runs), "failed runs to drop:\n")
  print(failed_runs)
} else {
  cat("\nExcellent: No runs fell below the 3 MAD threshold!\n")
}
```

## Bar plot of all Runs with MAD cut-off

```{r}
p1 <- ggplot(qc_data, aes(x = reorder(Run, Protein_Count), y = Protein_Count, fill = Condition)) +
  geom_bar(stat = "identity") +
  # ADD THE MAD CUTOFF LINE HERE
  geom_hline(yintercept = mad_cutoff, color = "red", linetype = "dashed", size = 1) +
  annotate("text", x = 5, y = mad_cutoff + 100, label = "3 MAD Cutoff", color = "red", hjust = 0) +
  theme_minimal() +
  coord_flip() + 
  scale_fill_manual(values = c("control" = "#1b9e77", "plaque_near" = "#d95f02", "plaque_far" = "#7570b3")) +
  labs(title = "Total Proteins Identified per Run",
       subtitle = "Red dashed line indicates the 3 MAD failure threshold",
       x = "Mass Spec Run", y = "Protein Count") +
  theme(axis.text.y = element_text(size = 4))

p1
```

## HeLa standards technical QC

The HeLa boxplots are level. This was consistent, showing mass-spec instrument stability.

```{r, fig.width=8, fig.height=5}
hela_cols <- grep("hela", quant_cols, value = TRUE, ignore.case = TRUE)
hela_log <- log2(as.matrix(spec_raw[, ..hela_cols]) + 1)
hela_log[hela_log == 0] <- NA 

# Plot HeLa Distributions
hela_long <- as.data.frame(hela_log) %>%
  pivot_longer(everything(), names_to = "Run", values_to = "Log2_Intensity") %>%
  mutate(Run_Short = str_extract(Run, "hela[0-9]+_S[0-9]+"))

p_hela <- ggplot(hela_long, aes(x = Run_Short, y = Log2_Intensity)) +
  geom_boxplot(fill = "lightblue", outlier.shape = NA) +
  theme_minimal() +
  labs(title = "HeLa QC Standards: Abundance Distribution",
       x = "HeLa Run", y = "Log2 Intensity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_hela)
```

## Abundance Distributions

Plotted log2intensities against each run (grouped by batch).

Batch 6 has a higher number of outliers than other batches.

Could be that these samples were thicker than others?

This deviation can be fixed by median normalisation.

```{r, fig.width=12, fig.height=6}
# Prepare full quantitative data
full_quant_log <- log2(as.matrix(quant_data) + 1)
full_quant_log[full_quant_log == 0] <- NA

# Merge with metadata and flag failures for the plot
abund_full <- as.data.frame(full_quant_log) %>%
  mutate(Protein = row_number()) %>%
  pivot_longer(-Protein, names_to = "Run", values_to = "Log2_Intensity") %>%
  drop_na() %>%
  left_join(qc_data, by = "Run") %>%
  mutate(Status = ifelse(Run %in% failed_runs, "Failed (Outlier)", "Pass"))

p_abund <- ggplot(abund_full, aes(x = reorder(Run, Batch), y = Log2_Intensity, fill = as.factor(Batch))) +
  geom_boxplot(aes(color = Status), outlier.shape = NA, lwd = 0.4) +
  scale_color_manual(values = c("Pass" = "black", "Failed (Outlier)" = "red")) +
  theme_minimal() +
  labs(title = "Log2 Abundance per Run (All Samples)",
       subtitle = "Red outlines indicate samples flagged by 3-MAD protein count filter",
       x = "Runs (Ordered by Batch)", y = "Log2 Intensity", fill = "Batch") +
  theme(axis.text.x = element_blank())

print(p_abund)
```

## Check correlations between samples

```{r, fig.width=10, fig.height=8}

#log2 transform matrix of intensities as pearson works best on log scale
quant_mat_log <- log2(as.matrix(quant_data) + 1)
quant_mat_log[quant_mat_log == 0] <- NA #ensure zeroes don't artificially boost correlation

#create correlation matrix
cor_mat <- cor(quant_mat_log, use = "pairwise.complete.obs")
# prepare the sorted metadata
# creating order for axes
meta_sorted <- meta_mapped %>%
  arrange(Condition, BioReplicate, Batch) %>%
  as.data.frame()

#Reorder the correlation matrix to match this biological order
# Run as index
sorted_runs <- meta_sorted$Run
cor_mat_sorted <- cor_mat[sorted_runs, sorted_runs]

#setup the annotation dataframe
anno_df <- meta_sorted %>%
  select(Run, Condition, BioReplicate, Batch) %>%
  mutate(Batch = as.factor(Batch)) %>%
  as.data.frame()

rownames(anno_df) <- anno_df$Run
anno_df$Run <- NULL

#Plot the Heatmap
pheatmap(cor_mat_sorted, 
         main = "Correlation Matrix: Grouped by Condition and Mouse",
         annotation_col = anno_df,
         cluster_rows = FALSE,  # DO NOT cluster - keep our manual order
         cluster_cols = FALSE,  # DO NOT cluster - keep our manual order
         show_rownames = FALSE, 
         show_colnames = FALSE,
         color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
         # Add gaps to visually separate the Conditions
         gaps_col = which(diff(as.numeric(as.factor(meta_sorted$Condition))) != 0),
         gaps_row = which(diff(as.numeric(as.factor(meta_sorted$Condition))) != 0))
```

## Identify outlier sample

Sample has 4 proteins total detected - below 3 MAD cut-off, so will be removed.

```{r}
# 1. Calculate Missingness and ID counts
qc_diagnostics <- data.table(
  Run = colnames(quant_data),
  Protein_Count = colSums(!is.na(quant_data) & quant_data > 0),
  NA_Percent = colMeans(is.na(quant_data) | quant_data == 0) * 100,
  Avg_Correlation = rowMeans(cor_mat, na.rm = TRUE)
)

# 2. Identify the outlier (lowest correlation)
outlier_id <- qc_diagnostics[which.min(Avg_Correlation), Run]

# 3. Print the diagnostic for the outlier
cat("--- Diagnostic for Heatmap Outlier ---\n")
print(qc_diagnostics[Run == outlier_id])
```

## Visualise outlier

```{r}
# 4. Visualize the relationship
p_diag <- ggplot(qc_diagnostics, aes(x = Protein_Count, y = Avg_Correlation)) +
  geom_point(aes(color = NA_Percent), size = 4) +
  geom_vline(xintercept = mad_cutoff, linetype = "dashed", color = "red") +
  geom_text(aes(label = ifelse(Run == outlier_id, "OUTLIER", "")), vjust = -1, color = "red") +
  scale_color_viridis_c(name = "% Missing") +
  theme_minimal() +
  labs(title = "Signal vs. Correlation Diagnostic",
       subtitle = "Red line is your 3-MAD cutoff. Does the outlier fall near it?",
       x = "Total Protein IDs", y = "Average Correlation (r)")

print(p_diag)
```

## Check for other samples which have lower correlations

All other samples seem to have decent correlations, which don't warrant removal.

```{r}
qc_data$Mean_Cor <- rowMeans(cor_mat, na.rm = TRUE)

# 2. Identify 'White Line' candidates 
# (Adjust the range if your scale's 'white' is centered differently, 
# but usually it's around 0.3 - 0.7 in this specific color map)
white_line_samples <- qc_data %>%
  filter(Mean_Cor < 0.7) %>%
  select(Run, Condition, Batch, Protein_Count, Mean_Cor)

# 3. Compare them to the 'Red' high-performers
cat("Summary of potential 'White Line' samples:\n")
print(white_line_samples)

# 4. Check for Batch bias in these lines
table(white_line_samples$Batch)
```

## Check for contaminants

Check for cumulative intensity of 10 most abundant proteins for each sample, and divide by total intensity for that run. Idcentifies low complexity samples where just a few proteins dominate signal. Again this shows that samples below the 3 MAD cut-off will be discarded.

```{r}
# 1. Ensure we have the raw intensities (non-log) for ratio calculations
# We use quant_data which was created from spec_raw earlier
raw_matrix <- as.matrix(quant_data)

# 2. Function to calculate % signal from Top N proteins
calc_top_n_pct <- function(column, n = 10) {
  sorted_vals <- sort(column, decreasing = TRUE, na.last = TRUE)
  top_n_sum <- sum(sorted_vals[1:n], na.rm = TRUE)
  total_sum <- sum(column, na.rm = TRUE)
  return((top_n_sum / total_sum) * 100)
}

# 2. Apply to all runs to create the metrics
top10_metrics <- apply(as.matrix(quant_data), 2, calc_top_n_pct, n = 10)

# 3. Add the missing column to qc_data
qc_data[, Top10_Percent := top10_metrics[Run]]

# 4. Identify suspect runs (Threshold > 50%)
contaminant_suspects <- qc_data[Top10_Percent > 50, .(Run, Condition, Batch, Protein_Count, Top10_Percent)]

cat("Samples where the Top 10 proteins account for >50% of signal:\n")
print(contaminant_suspects)
```

## Check for what specific proteins are contributing to these mid zone proteins

Gemini output of these protein:

-   **A2AQ07**: Plectin (Mouse) – A giant structural protein that links the cytoskeleton to the plasma membrane. It is extremely common in dense tissue cuts.

-   **P02535**: Keratin, type I cytoskeletal 10 (Mouse).

-   **P60710**: Actin, cytoplasmic 1 (Mouse).

-   **P68372**: Tubulin beta-4B chain (Mouse).

-   **Q5U405**: Transmembrane protease serine 13 (Mouse).

-   **Q9QWL7**: Keratin, type I cytoskeletal 17 (Mouse).

```{r}
# 1. Calculate Total Raw Intensity per Run
# We use the raw quant_data (pre-log) to see the physical signal magnitude
total_intensities <- colSums(quant_data, na.rm = TRUE)

# 2. Add Total_Intensity to your existing qc_data
# Match the names of the colSums output to the Run column in qc_data
qc_data[, Total_Intensity := total_intensities[Run]]

# 3. Create the Diagnostic Plot
p_intensity_vs_count <- ggplot(qc_data, aes(x = Protein_Count, y = Total_Intensity)) + 
  # Use log10 scale for Y if intensities vary by orders of magnitude
  scale_y_log10() + 
  geom_point(aes(color = as.factor(Batch)), size = 3, alpha = 0.7) +
  # Add the 3-MAD failure line for context
  geom_vline(xintercept = mad_cutoff, linetype = "dashed", color = "red") +
  # Label the outlier specifically
  geom_text(aes(label = ifelse(Run == outlier_id, "Outlier", "")), 
            vjust = -1, color = "darkred", fontface = "bold") +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Diagnostic: Protein IDs vs. Total Signal",
       subtitle = "Checking for high-intensity contaminants or low-yield runs",
       x = "Total Proteins IDs",
       y = "Total Intensity (Log10 Scale)",
       color = "Batch")

print(p_intensity_vs_count)
```

## Top 10 protein signal contribution against total number of proteins identified.

```{r}
p_top10 <- ggplot(qc_data, aes(x = Protein_Count, y = Top10_Percent)) +
  geom_point(aes(color = as.factor(Batch), size = Total_Intensity), alpha = 0.6) +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red") +
  # Label samples that cross the 50% threshold
  geom_text(aes(label = ifelse(Top10_Percent > 50, Run, "")), 
            vjust = -1, size = 3, check_overlap = TRUE) +
  theme_minimal() +
  labs(title = "Top 10 Protein Signal Contribution",
       subtitle = "Above 50% indicates potential contamination or extremely low complexity",
       x = "Total Proteins Identified",
       y = "% Total Intensity from Top 10 Proteins",
       size = "Total Signal")

print(p_top10)
```

```{r}
mid_zone_runs <- qc_data[Top10_Percent > 30 & Top10_Percent < 50, Run]

# Get the top protein ID for each of these runs
top_id_check <- apply(quant_data[, ..mid_zone_runs], 2, function(x) {
  row_idx <- which.max(x)
  # spec_raw[[1]] pulls the first column (usually PG.ProteinGroups or similar)
  return(spec_raw[[1]][row_idx]) 
})

print(table(top_id_check))
```

## Filter out samples below 3 MAD cut-off

```{r}
final_runs <- setdiff(valid_runs, failed_runs)
cat("Original runs:", length(valid_runs), "\n")
cat("Runs dropped:", length(failed_runs), "\n")
cat("Clean runs remaining:", length(final_runs), "\n\n")


#Subset raw matrix to keep only clean runs
clean_quant <- as.matrix(quant_data[, ..final_runs])
```

# Save objects

Save filtered metadata and cleaned raw Spectronaut output

```{r}
#metadata save
clean_meta <- qc_data[Run %in% final_runs]

#Format filtered spectronaut data
anno_cols <- c("PG.ProteinGroups", "PG.Genes", "PG.ProteinDescriptions")
all_desired_cols <- c(anno_cols, final_runs)
clean_spec_raw <- spec_raw[, ..all_desired_cols]
head(clean_spec_raw)

#Format dropped samples
dropped_log <- qc_data[Run %in% failed_runs, .(Run, Condition, Batch, Protein_Count)]
dropped_log[, Reason := "Sub 3-MAD Protein Count"]
dropped_log

# Save objects
fwrite(clean_meta, file.path(objects_dir, "clean_metadata.csv"))
fwrite(clean_spec_raw, file.path(objects_dir, "clean_spec_raw.csv"))
fwrite(dropped_log, file.path(objects_dir, "dropped_samples_log.csv"))
```
