---
title: "ms_stats_initial"
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics')
library(data.table)
library(MSstats)
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
base_dir <- "/Volumes/lab-destrooperb/home/shared/zanettc/millie_proteomics"

run_num <- "run1"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in MS stats spectronaut output

```{r}
#raw_spectronaut <- fread(file.path(raw_dir, "20260218_092804_hela20_Report_msstats.tsv"), sep="\t", header=TRUE)
```

## Save as .qs2 object to save space

```{r}
#qs_save(raw_spectronaut,file.path(objects_dir, "msstats_report.qs2"))
```

## Read in .qs2 object

```{r}
raw_spectronaut <- qs_read(file.path(objects_dir, "msstats_report.qs2"))
```

## Read in metadata

```{r}
clean_meta <- fread(file.path(objects_dir, "clean_metadata.csv"))
```

```{r}
msstats_mapped <- raw_spectronaut %>%
  mutate(
    # Extract the well/chip location
    Chip_location = str_match(R.FileName, "-([A-Z][0-9]+)_")[, 2],
    
    # Extract the raw batch
    Batch_raw = str_match(tolower(R.FileName), "mt_[a-z0-9]+_([0-9]+|s3)")[, 2],
    
    # Fix the "s3" batch 5 naming error and make numeric
    Batch = as.numeric(str_replace(Batch_raw, "s3", "5"))
  ) %>%
  select(-Batch_raw)

length(unique(msstats_mapped$R.FileName))

```

```{r}
msstats_clean <- msstats_mapped %>%
  inner_join(
    # Select only the columns needed for mapping to avoid duplicating metadata
    clean_meta %>% select(Run_wide = Run, Chip_location, Batch, Condition), 
    by = c("Chip_location", "Batch")
  ) %>%
  # Overwrite the mismatched MSstats Run name with your standardized wide Run name
  mutate(Run = Run_wide) %>%
  select(-Run_wide)

length(unique(msstats_clean$R.FileName))

#Prepare the mapping from metadata - select unique identifiers
meta_mapping_keys <- clean_meta %>% 
  select(Chip_location, Batch, Condition, BioReplicate)

# Update MSstats data and use its Run name as the primary ID
msstats_clean <- msstats_mapped %>%
  # Join to only keep the runs that passed your initial wide-format QC
  inner_join(meta_mapping_keys, by = c("Chip_location", "Batch")) %>%
  as.data.table()

# 3. Overwrite the Metadata Run names to match the MSstats filenames
final_clean_meta <- msstats_clean %>%
  select(R.FileName, Chip_location, Batch, Condition, BioReplicate) %>%
  distinct() %>%
  as.data.table()

head(msstats_clean)

msstats_clean

final_clean_meta
cat("Original MSstats rows:", nrow(raw_spectronaut), "\n")
cat("Cleaned MSstats rows:", nrow(msstats_clean), "\n")

# Save the filtered and aligned MSstats input for downstream steps
qs_save(msstats_clean, file.path(objects_dir, "clean_msstats_input.qs2"))
```

## Aß tryptic peptide

```{r}
# Define the target peptide based on your mouse model
target_peptide <- "LVFFAEDVGSNK"
colnames(msstats_clean)
# Filter the cleaned MSstats data for the Aß peptide
ab_check_data <- msstats_clean %>%
  filter(PEP.StrippedSequence == target_peptide) %>%
  # Use F.PeakArea or F.NormalizedPeakArea for the raw intensity values
  group_by(R.FileName, Condition, Batch) %>%
  summarise(
    Total_Intensity = sum(F.PeakArea, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Log2_Intensity = log2(Total_Intensity + 1))

# Generate the facet plot to expose the batch swap
p_ab_check <- ggplot(ab_check_data, aes(x = Condition, y = Log2_Intensity, fill = Condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.8) +
  facet_wrap(~ Batch, scales = "free_y") + 
  theme_minimal() +
  scale_fill_manual(values = c("control" = "#1b9e77", "plaque_near" = "#d95f02", "plaque_far" = "#7570b3")) +
  labs(
    title = "Amyloidome Verification: Aβ Tryptic Peptide",
    subtitle = "Confirming sample swaps: expect high signal in 'plaque_near' [cite: 54, 203]",
    x = "Labelled Condition",
    y = "Log2(Aβ Peptide Intensity)"
  )

# Save the plot
ggsave(file.path(results_dir, "AB_Peptide_Batch_Verification.pdf"), p_ab_check, width = 10, height = 6)
print(p_ab_check)
```

## Best option now is to continue analysis with just 1,3,4 

Extremely hard to identify and map back to original data.

```{r}
good_batches <- c(1, 3, 4)

#Filter the metadata
final_clean_meta <- clean_meta %>%
  filter(Batch %in% good_batches) %>%
  as.data.table()

msstats_annotation <- final_clean_meta %>%
  mutate(Run = str_replace_all(Run, 
                               # 1. Matches leading "[101] "
                               # 2. Matches trailing ".d.PG.Quantity"
                               pattern = "^\\[[0-9]+\\]\\s*|\\.d\\.PG\\.Quantity$", 
                               replacement = "")) %>%
  select(Run, Condition, BioReplicate) %>%
  distinct()

length(msstats_annotation$Run)

if (all(sample_meta %in% unique(final_msstats_input$R.FileName))) {
  cat("✅ Alignment Success: Metadata now matches MSstats R.FileName.\n")
} else {
  stop("❌ Alignment Failure")
}









# Filter the MSstats input data
final_msstats_input <- msstats_clean %>%
  filter(Batch %in% good_batches) %>%
  as.data.table()

cat("Remaining pristine samples in metadata:", nrow(final_clean_meta), "\n")
cat("Remaining rows in MSstats data:", nrow(final_msstats_input), "\n")

fwrite(final_clean_meta, file.path(objects_dir, "final_clean_metadata.csv"))
frwrite(msstats_annotation, file.path(objects_dir, "msstats_annotation.csv"))
qs_save(final_msstats_input, file.path(objects_dir, "final_msstats_input.qs2"))
```

## Format conversion

```{r}
msstats_input <- SpectronauttoMSstatsFormat(
  final_msstats_input,
  annotation = final_clean_meta,
  filter_with_Qvalue = TRUE,    # Keeps only high-confidence identifications
  qvalue_cutoff = 0.01,         # 1% FDR
  fewMeasurements = "remove",   # Drops peptides with too few data points to model
  removeProtein_with1Feature = FALSE # Keep single-peptide proteins if desired
)
```

```{r}
processed_data <- dataProcess(
  msstats_input,
  normalization = "equalizeMedians", # Matches the paper's median intensity normalization
  summaryMethod = "TMP",             # Robust summarization
  censoredInt = "NA",                # Spectronaut outputs NA for missing/below LOD
  MBimpute = FALSE,                  # IMPROVEMENT: Do not blindly impute missing values to avoid over-interpretation
  featureSubset = "highQuality",     # IMPROVEMENT: Drop noisy peptides before summarizing
  remove_uninformative_feature_outlier = TRUE # Helps clean up LCM noise
)

# Save the processed object 
qs_save(processed_data, file.path(objects_dir, "processed_msstats_data.qs2"))
```

```{r}
processed_data <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))
```

```{r}
# 1. Verify the alphabetical order of your conditions
# It should be: "control", "plaque_far", "plaque_near"
condition_levels <- levels(processed_data$ProteinLevelData$GROUP)
print(condition_levels) 

# 2. Build the contrast matrix
# Rows are comparisons, columns correspond to condition_levels
comparison_matrix <- matrix(c(
  -1,  0,  1,  # PlaqueNear vs Control (PlaqueNear - Control)
  -1,  1,  0,  # PlaqueFar vs Control (PlaqueFar - Control)
   0, -1,  1   # PlaqueNear vs PlaqueFar (PlaqueNear - PlaqueFar)
), nrow = 3, byrow = TRUE)

rownames(comparison_matrix) <- c("PlaqueNear_vs_Control", "PlaqueFar_vs_Control", "PlaqueNear_vs_PlaqueFar")
colnames(comparison_matrix) <- condition_levels

# 3. Run the statistical modeling
test_results <- groupComparison(contrast.matrix = comparison_matrix, data = processed_data)

# Extract the results table
res_df <- test_results$ComparisonResult

# Save the full and filtered results
fwrite(res_df, file.path(results_dir, "Full_GroupComparison_Results.csv"))
```





## correlate between Aß peptide and every other protein in dataset 
looking for ptoeins highly correlated
```{r}

```




## go gsea, MEGENA?
```{r}

```

#correlate with single plaque paper? which proteins are new?
