---
title: "2a. MSstats workflow"
---

## Libraries

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics')
library(data.table)
library(MSstats)
options(MSstatsLog = "ERROR")
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(stringr)
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run5"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in MS stats spectronaut output

```{r}
#Raw TSV import - commented out as using .qs2 format saved below
#raw_spectronaut <- fread(file.path(raw_dir, "20260218_092804_hela20_Report_msstats.tsv"), sep="\t", header=TRUE)
```

## Save as .qs2 object to save space

```{r}
#qs_save(raw_spectronaut,file.path(objects_dir, "msstats_report.qs2"))
```

## Read in .qs2 object

```{r}
raw_spectronaut <- qs_read(file.path(base_dir, "data/processed/run1/msstats_report.qs2"))
raw_spec
```

## Read in metadata

```{r}
clean_meta <- fread(file.path(base_dir, "data/processed/run1/clean_metadata.csv"))
```

## Map msstats filenames to Metadata

```{r}
msstats_mapped <- raw_spectronaut %>%
  mutate(
    # Extract the well/chip location
    Chip_location = str_match(R.FileName, "-([A-Z][0-9]+)_")[, 2],
    
    # Extract the raw batch
    Batch_raw = str_match(tolower(R.FileName), "mt_[a-z0-9]+_([0-9]+|s3)")[, 2],
    
    # Fix the "s3" batch 5 naming error and make numeric
    Batch = as.numeric(str_replace(Batch_raw, "s3", "5"))
  ) %>%
  select(-Batch_raw)
```

## Joining raw data with metadata

```{r}
msstats_clean <- msstats_mapped %>%
  inner_join(
    clean_meta %>% select(Run_wide = Run, Chip_location, Batch, Condition, BioReplicate), 
    by = c("Chip_location", "Batch")
  ) %>%
  # Overwrite the mismatched MSstats Run name with your standardized wide Run name
  mutate(Run = Run_wide) %>%
  select(-Run_wide) %>%
  as.data.table()

# 3. Overwrite the Metadata Run names to match the MSstats filenames
final_clean_meta <- msstats_clean %>%
  select(R.FileName, Chip_location, Batch, Condition, BioReplicate) %>%
  distinct() %>%
  as.data.table()

cat("Original MSstats rows:", nrow(raw_spectronaut), "\n")
cat("Cleaned MSstats rows:", nrow(msstats_clean), "\n")

# Save the filtered and aligned MSstats input for downstream steps
#qs_save(msstats_clean, file.path(objects_dir, "clean_msstats_input.qs2"))
```

## Aß tryptic peptide check

```{r}
# Define the target peptide based on your mouse model
target_peptide <- "LVFFAEDVGSNK"
colnames(msstats_clean)
# Filter the cleaned MSstats data for the Aß peptide
ab_check_data <- msstats_clean %>%
  filter(PEP.StrippedSequence == target_peptide) %>%
  # Use F.PeakArea or F.NormalizedPeakArea for the raw intensity values
  group_by(R.FileName, Condition, Batch) %>%
  summarise(
    Total_Intensity = sum(F.PeakArea, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Log2_Intensity = log2(Total_Intensity + 1))

# Generate the facet plot to expose the batch swap
p_ab_check <- ggplot(ab_check_data, aes(x = Condition, y = Log2_Intensity, fill = Condition)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.8) +
  facet_wrap(~ Batch, scales = "free_y") + 
  theme_minimal() +
  scale_fill_manual(values = c("control" = "#1b9e77", "plaque_near" = "#d95f02", "plaque_far" = "#7570b3")) +
  labs(
    title = "Amyloidome Verification: Aβ Tryptic Peptide",
    subtitle = "Confirming sample swaps: expect high signal in 'plaque_near' [cite: 54, 203]",
    x = "Labelled Condition",
    y = "Log2(Aβ Peptide Intensity)"
  )

# Save the plot
ggsave(file.path(results_dir, "AB_Peptide_Batch_Verification.pdf"), p_ab_check, width = 10, height = 6)
print(p_ab_check)
```

## Best option now is to continue analysis with just 1,3,4

```{r}
#Filter only for good batches
good_batches <- c(1, 3, 4)

#Filter the metadata
final_msstats_input <- msstats_clean %>%
  filter(Batch %in% good_batches) %>%
  as.data.table()

#Build annotation metrix MSstats needs from filtered data
#MSstats requires Run, Condition, BioReplicate
msstats_annotation <- final_msstats_input %>%
  mutate(Run = str_replace_all(Run, 
                               # 1. Matches leading "[101] "
                               # 2. Matches trailing ".d.PG.Quantity"
                               pattern = "^\\[[0-9]+\\]\\s*|\\.d\\.PG\\.Quantity$", 
                               replacement = "")) %>%
  select(Run, Condition, BioReplicate) %>%
  distinct()

cat("Remaining pristine samples in metadata:", nrow(msstats_annotation), "\n")
cat("Remaining rows in MSstats data:", nrow(final_msstats_input), "\n")

#save final input files so don't have to re-run cleaning steps
qs_save(final_msstats_input, file.path(objects_dir, "final_msstats_input.qs2"))
fwrite(msstats_annotation, file.path(objects_dir, "msstats_annotation.csv"))


```


##Rerun - load in data and edit to remove Keratin
```{r}
final_msstats_input <- qs_read(file.path(base_dir, "data/processed/run1/final_msstats_input.qs2"))
msstats_annotation <- fread(file.path(base_dir, "data/processed/run1/msstats_annotation.csv"))


final_msstats_input <- final_msstats_input %>%
  filter(!str_detect(PG.ProteinGroups, "^(?i)Krt")) %>%
  as.data.table()
final_msstats_input
```

```{r}
#Initial feature-level filtering and maps Spectronaut columns to MSstats required headers
msstats_input <- SpectronauttoMSstatsFormat(
  final_msstats_input,
  annotation = msstats_annotation,
  filter_with_Qvalue = TRUE,    # Keeps only high-confidence identifications
  qvalue_cutoff = 0.01,         # 1% FDR
  fewMeasurements = "remove",   # Drops peptides with too few data points to model
  removeProtein_with1Feature = FALSE # Keep single-peptide proteins if desired
)
qs_save(msstats_input, file.path(objects_dir, "msstats_input.qs2"))
```

## input filtering at the condition level
```{r}
msstats_input <- qs_read(file.path(base_dir, "data/processed/run2/msstats_input.qs2"))
# 1. Convert to a standard data frame
input_df <- as.data.frame(msstats_input)

# 2. Set threshold
threshold <- 0.5

# 3. Filter and join with the extra "Condition" key to prevent many-to-many errors
msstats_input_filtered <- input_df %>%
  group_by(ProteinName, PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge, Condition) %>%
  summarize(
    n_obs = sum(!is.na(Intensity) & Intensity > 0),
    n_total = n(),
    prop = n_obs / n_total,
    .groups = "drop"
  ) %>%
  # Keep features that meet the 30% threshold in AT LEAST one condition
  group_by(ProteinName, PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge) %>%
  filter(any(prop >= threshold)) %>%
  ungroup() %>%
  # Join back using ALL keys including Condition to ensure 1-to-1 matching
  inner_join(input_df, by = c("ProteinName", "PeptideSequence", "PrecursorCharge", 
                              "FragmentIon", "ProductCharge", "Condition"))

# Diagnostic
message("Original Rows: ", nrow(input_df))
message("Filtered Rows: ", nrow(msstats_input_filtered))

```

## Filtering out keratin
```{r}
clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))

protein_dictionary <- clean_spec_raw %>%
  dplyr::select(Protein = PG.ProteinGroups, 
         Gene = PG.Genes, 
         Description = PG.ProteinDescriptions) %>%
  distinct()

msstats_input_filtered_keratin <- msstats_input_filtered %>%
  left_join(protein_dictionary, by=c("ProteinName" = "Protein")) %>%
  filter(!grepl("^Krt", Gene)) %>%
  dplyr::select(-c("Gene", "Description"))
  
msstats_input_filtered_keratin
```
## Data summarisation and normalisation

```{r}
# Roll up the peptide intensities to protein-level abundances

processed_data <- dataProcess(
  msstats_input_filtered_keratin,
  normalization = "equalizeMedians",
  summaryMethod = "TMP",
  censoredInt = "NA",
  MBimpute = FALSE,
  featureSubset = "highQuality",
  remove_uninformative_feature_outlier = TRUE
  )

# Save the processed object 
qs_save(processed_data, file.path(objects_dir, "processed_msstats_data.qs2"))
```

## Reload data in if needed

```{r}
processed_data <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))
```

## Statistical modelling

```{r}
# 1. Verify the alphabetical order of your conditions
# It should be: "control", "plaque_far", "plaque_near"
condition_levels <- levels(processed_data$ProteinLevelData$GROUP)
print(condition_levels) 

# 2. Build the contrast matrix
# Rows are comparisons, columns correspond to condition_levels
comparison_matrix <- matrix(c(
  -1,  0,  1,  # PlaqueNear vs Control (PlaqueNear - Control)
  -1,  1,  0,  # PlaqueFar vs Control (PlaqueFar - Control)
   0, -1,  1   # PlaqueNear vs PlaqueFar (PlaqueNear - PlaqueFar)
), nrow = 3, byrow = TRUE)

rownames(comparison_matrix) <- c("PlaqueNear_vs_Control", "PlaqueFar_vs_Control", "PlaqueNear_vs_PlaqueFar")
colnames(comparison_matrix) <- condition_levels

# Run linear mixed-effects model across all proteins
test_results <- groupComparison(contrast.matrix = comparison_matrix, data = processed_data)

qs_save(test_results, file.path(objects_dir, "comparison_results.qs2"))

# Extract the results table
res_df <- test_results$ComparisonResult

# Save the full and filtered results
fwrite(res_df, file.path(results_dir, "Full_GroupComparison_Results.csv"))
```
