---
title: "2b. MSstats plots"
format: html
editor: visual
---

# Model-based analysis plots

Linear mixed effect models make two big assumptions:

1.  Errors are normally distributed
2.  Variance is constant

Ensures that significant proteins aren't statistical artifacts

Generated custom plots as have gene names + proteins labelled here for easy checking

## Libraries

```{r}
library(data.table)
library(MSstats)
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(glue)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run5"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")
millie_dir <- file.path(base_dir, "data", "processed_millie")


dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

# Read in processed MSstats results

```{r}
msstats_results <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))
msstats_comparison <- qs_read(file.path(objects_dir, "comparison_results.qs2"))
millie_combined <- fread(file.path(millie_dir, "millie_combined.csv"))
```

## Map protein IDs to genes

```{r}
clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))

protein_dictionary <- clean_spec_raw %>%
  dplyr::select(PROTEIN = PG.ProteinGroups, 
         Gene = PG.Genes, 
         Description = PG.ProteinDescriptions) %>%
  distinct()

# Update feature level data for profile plots
msstats_results$FeatureLevelData <- msstats_results$FeatureLevelData %>%
  left_join(protein_dictionary, by = "PROTEIN") %>%
  mutate(PROTEIN = ifelse(!is.na(Gene), paste(Gene, PROTEIN, sep = "_"), PROTEIN)) %>%
  dplyr::select(-Gene)



#Update ProteinLevelData -> necessary for condition plots
msstats_results$ProteinLevelData <- msstats_results$ProteinLevelData %>%
  left_join(protein_dictionary, by = c("Protein" = "PROTEIN")) %>%
  mutate(Protein = ifelse(!is.na(Gene) & Gene != "", 
                          paste(Gene, Protein, sep = "_"), 
                          Protein)) %>%
  dplyr::select(-Gene)


#update comparison object  

msstats_comparison$ComparisonResult <- msstats_comparison$ComparisonResult %>%
  mutate(Protein = as.character(Protein))%>% # Convert to character to join/edit
  left_join(protein_dictionary, by = c("Protein" = "PROTEIN")) %>%
  mutate(Protein = ifelse(!is.na(Gene), 
                          paste(Gene, Protein, sep = "_"), 
                          Protein)) %>%
  dplyr::select(-Gene, -Description) %>%
  mutate(Protein = factor(Protein))

#update ModelQC object
msstats_comparison$ModelQC <- msstats_comparison$ModelQC %>%
  mutate(Protein = as.character(Protein)) %>% # Convert to character to join/edit
  left_join(protein_dictionary, by = c("Protein" = "PROTEIN")) %>%
  mutate(Protein = ifelse(!is.na(Gene) & Gene != "", 
                          paste(Gene, Protein, sep = "_"), 
                          Protein)) %>%
  dplyr::select(-Gene, -Description) %>%
  mutate(Protein = factor(Protein))
```

## QC plots for all proteins

```{r}
msstats_dir <- file.path(results_dir, "msstats_plots")
dir.create(msstats_dir, recursive = TRUE, showWarnings = FALSE)

# Condition Plot: Visualizes the mean and 95% CI per condition
target_ids <- protein_dictionary %>%
  mutate(Combined = paste(Gene, PROTEIN, sep = "_")) %>%
  pull(Combined)

pdf_path <- file.path(msstats_dir, "ConditionPlot.pdf")
pdf(pdf_path, width = 10, height = 7)

# Loop through specific target IDs
for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping ConditionPlot for: ", prot, " - Not found in msstats_results.")
    next 
  }

  dataProcessPlots(data = msstats_results,
                   type = "Conditionplot",
                   which.Protein = prot, # This MUST be a single string from your vector
                   address = FALSE)      # Plots to the open 'pdf' device
}
dev.off()

# QC plot

pdf_path <- file.path(msstats_dir, "qc_summaries.pdf")
pdf(pdf_path, width = 10, height = 7)
for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping ConditionPlot for: ", prot, " - Not found in msstats_results.")
    next 
  }
  dataProcessPlots(data = msstats_results,
                   type = "QCplot",
                   which.Protein = prot,
                   address = FALSE)
}
dev.off()

#Profile Plot: Shows individual feature trends
pdf_path <- file.path(msstats_dir, "profile_features.pdf")
pdf(pdf_path, width = 10, height = 7)

for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping ConditionPlot for: ", prot, " - Not found in msstats_results.")
    next 
  }
  dataProcessPlots(data = msstats_results,
                   type = "Profileplot",
                   which.Protein = prot,
                   address = FALSE)
  }
dev.off()
  
#Profile Plot: Shows individual peptide trends
pdf_path <- file.path(msstats_dir, "profile_peptides.pdf")

pdf(pdf_path, width = 10, height = 7)

for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping ConditionPlot for: ", prot, " - Not found in msstats_results.")
    next 
  }
  dataProcessPlots(data = msstats_results,
                   type = "Profileplot",
                   featureName = "peptide",
                   which.Protein = prot,
                   address = FALSE)
  }
dev.off()

```


## Residual plot
```{r}
pdf_path_res <- file.path(msstats_dir, "custom_residual_plots.pdf")
pdf(pdf_path_res, width = 10, height = 7)

for (prot in target_ids) {
  # Filter data for the current protein
  prot_data <- msstats_comparison$ModelQC %>% 
    filter(Protein == prot)
  
  # Safety check: skip to the next protein if no data is found
  if (nrow(prot_data) == 0) {
    message("No ModelQC data found for: ", prot, " - Skipping.")
    next
  }
  
  message("Plotting Custom Residuals for: ", prot)
  
  # Build the residual plot
  p_res <- ggplot(prot_data, aes(x = fitted, y = residuals)) +
    geom_point(alpha = 0.6, size = 2.5, color = "dimgrey") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "darkred", linewidth = 0.8) +
    labs(title = paste("Residual Plot:", prot), 
         x = "Predicted Abundance (Fitted)", 
         y = "Residuals") +
    theme_bw() +
    theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  
  # print to the open PDF device
  print(p_res)
}
dev.off()
```

## QQ plots

Checks normal distribution

```{r}
pdf_path_qq <- file.path(msstats_dir, "custom_qq_plots.pdf")
pdf(pdf_path_qq, width = 10, height = 7)

for (prot in target_ids) {
  # Filter data for the current protein
  prot_data <- msstats_comparison$ModelQC %>% 
    filter(Protein == prot)
  #safety check
  if (nrow(prot_data) == 0) next
  
  message("Plotting Custom QQ for: ", prot)
  
  # Build the qq plot
  p_qq <- ggplot(prot_data, aes(sample = residuals)) +
    stat_qq(alpha = 0.6, size = 2.5, color = "dimgrey") +
    stat_qq_line(color = "dodgerblue", linewidth = 0.8) +
    labs(title = paste("QQ Plot:", prot),
         x = "Theoretical Quantiles",
         y = "Sample Quantiles") +
    theme_bw() +
    theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  
  # print to the open PDF device
  print(p_qq)
}
dev.off()

```


## Get significant Millie genes for Plaque Near vs Far

```{r}
sig_limma_genes <- millie_combined %>%
  filter(Label == "PlaqueNear_vs_PlaqueFar" & Status_limma != "Not Significant") %>%
  pull(Gene)

target_ids <- protein_dictionary %>%
  filter(Gene %in% sig_limma_genes) %>%
  mutate(Combined = paste(Gene, PROTEIN, sep = "_")) %>%
  pull(Combined)

```

# Visualisation of limma processed data

### QC profile and condition plots

```{r}
# Condition Plot: Visualizes the mean and 95% CI per condition
limma_path <- file.path(results_dir, "msstats_plots", "limma")
dir.create(limma_path, recursive = TRUE, showWarnings = FALSE)

pdf_path <- file.path(results_dir, "msstats_plots",  "limma", "ConditionPlot.pdf")
pdf(pdf_path, width = 10, height = 7)

# Loop through your specific target IDs
for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping ConditionPlot for: ", prot, " - Not found in msstats_results.")
    next 
  }

  dataProcessPlots(data = msstats_results,
                   type = "Conditionplot",
                   which.Protein = prot, # This MUST be a single string from your vector
                   address = FALSE)      # Plots to the open 'pdf' device
}
dev.off()

# QC plot

pdf_path <- file.path(results_dir, "msstats_plots",  "limma", "qc_summaries.pdf")
pdf(pdf_path, width = 10, height = 7)
for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping QCPlot for: ", prot, " - Not found in msstats_results.")
    next 
  }
  dataProcessPlots(data = msstats_results,
                   type = "QCplot",
                   which.Protein = prot,
                   address = FALSE)
}
dev.off()

#Profile Plot: Shows individual feature trends
pdf_path <- file.path(results_dir, "msstats_plots",  "limma", "profile_features.pdf")
pdf(pdf_path, width = 10, height = 7)

for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping FeaturePlot for: ", prot, " - Not found in msstats_results.")
    next 
  }
  dataProcessPlots(data = msstats_results,
                   type = "Profileplot",
                   which.Protein = prot,
                   address = FALSE)
  }
dev.off()
  
#Profile Plot: Shows individual peptide trends
pdf_path <- file.path(results_dir, "msstats_plots",  "limma", "profile_peptides.pdf")

pdf(pdf_path, width = 10, height = 7)

for (prot in target_ids) {
  if (!(prot %in% msstats_results$ProteinLevelData$Protein)) {
    message("Skipping PeptidePlot for: ", prot, " - Not found in msstats_results.")
    next 
  }
  dataProcessPlots(data = msstats_results,
                   type = "Profileplot",
                   featureName = "peptide",
                   which.Protein = prot,
                   address = FALSE)
  }
dev.off()
```

# Model-based analysis plots

Linear mixed effect models make two big assumptions:

1.  Errors are normally distributed
2.  Variance is constant

Ensures that significant proteins aren't statistical artifacts

Generated custom plots as have gene names + proteins labelled here for easy checking

## Generate custom residual plots

Residuals should be random

```{r}
pdf_path_res <- file.path(results_dir, "msstats_plots", "limma", "custom_residual_plots.pdf")
pdf(pdf_path_res, width = 10, height = 7)

for (prot in target_ids) {
  # Filter data for the current protein
  prot_data <- msstats_comparison$ModelQC %>% 
    filter(Protein == prot)
  
  # Safety check: skip to the next protein if no data is found
  if (nrow(prot_data) == 0) {
    message("No ModelQC data found for: ", prot, " - Skipping.")
    next
  }
  
  message("Plotting Custom Residuals for: ", prot)
  
  # Build the residual plot
  p_res <- ggplot(prot_data, aes(x = fitted, y = residuals)) +
    geom_point(alpha = 0.6, size = 2.5, color = "dimgrey") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "darkred", linewidth = 0.8) +
    labs(title = paste("Residual Plot:", prot), 
         x = "Predicted Abundance (Fitted)", 
         y = "Residuals") +
    theme_bw() +
    theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  
  # print to the open PDF device
  print(p_res)
}
dev.off()
```

## QQ plots

Checks normal distribution

```{r}
pdf_path_qq <- file.path(results_dir, "msstats_plots", "limma", "custom_qq_plots.pdf")
pdf(pdf_path_qq, width = 10, height = 7)

for (prot in target_ids) {
  # Filter data for the current protein
  prot_data <- msstats_comparison$ModelQC %>% 
    filter(Protein == prot)
  #safety check
  if (nrow(prot_data) == 0) next
  
  message("Plotting Custom QQ for: ", prot)
  
  # Build the qq plot
  p_qq <- ggplot(prot_data, aes(sample = residuals)) +
    stat_qq(alpha = 0.6, size = 2.5, color = "dimgrey") +
    stat_qq_line(color = "dodgerblue", linewidth = 0.8) +
    labs(title = paste("QQ Plot:", prot),
         x = "Theoretical Quantiles",
         y = "Sample Quantiles") +
    theme_bw() +
    theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5))
  
  # print to the open PDF device
  print(p_qq)
}
dev.off()

```
