---
title: "06: MSstats limma correlation"
format: html
---

==================================================================

# MSstats vs limma Concordance Analysis

Author: Carlo Zanetti

This script evaluates the consistency between MSstats and limma. It merges results from both methods to identify shared biological signals and pipeline specific differences.

### Workflow overview

1.  **Data loading and visualisation**

    Import Msstats and limma data

2.  **Gene mapping and Annotation**

    Only necessary for Msstats

3.  **Thresholding and categorisation**

    Apply significance cut-offs (FDR \<0.05, \|log2FC\| \> 0.5)

4.  **Concordance analysis**

    Calculate pearson correlation between log2FC values of the 2 pipelines for each comparison

5.  **Visualisation**

    Plots concordance scatter plots between each pipeline and comparison.

    Labels top 30 most highly changed genes using ggrepel

    Produces tables summarising detction overlap and pipeline-specific abundant proteins

==================================================================

## Load libraries

```{r}
library(data.table)
library(qs2)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(purrr)
library(stringr)
```

## Read in limma and msstats objects

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run3"
results_dir <- file.path(base_dir, "results", run_num)

objects_dir <- file.path(base_dir, "data", "processed", run_num)
millie_dir <- file.path(base_dir, "data", "processed_millie")
graphs_dir <- file.path(results_dir, "concordance")


millie_f_v_c <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_far_vs_Control.csv"))
millie_n_v_c <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_near_vs_Control.csv"))
millie_n_v_f <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_near_vs_Far.csv"))

res_df <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))
clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))

```

## Convert msstats to gene names

```{r}
protein_dictionary <- clean_spec_raw %>%
  select(Protein = PG.ProteinGroups, 
         Gene = PG.Genes, 
         Description = PG.ProteinDescriptions) %>%
  distinct()

msstats_ready <- res_df %>%
  # Filter for your specific comparison (Label) and handle NAs
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  
  # Join with dictionary
  left_join(protein_dictionary, by = "Protein") %>%
  
  # Clean up Gene names and define significance
  mutate(
    # Fallback to Protein ID if Gene name is missing or empty
    Gene = ifelse(is.na(Gene) | Gene == "", Protein, Gene),
    
    # Define significance categories
    Status = case_when(
      adj.pvalue < 0.05 & log2FC > 0.5 ~ "Upregulated",
      adj.pvalue < 0.05 & log2FC < -0.5 ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  ) %>%
  
  # Ranking for plotting/reporting
  group_by(Label, Status) %>%
  mutate(Rank = rank(adj.pvalue, ties.method = "first")) %>%
  ungroup()


msstats_ready %>% 
  filter(Label == "PlaqueNear_vs_PlaqueFar") %>%
  filter(log2FC > 0.5 & adj.pvalue < 0.05)
```

## Combine millie's dataframes into a mega merged one

```{r}
# Add labels to Millie's data to match MSstats 'Label' names
# Check your msstats_ready$Label to ensure these strings match exactly!
millie_combined <- bind_rows(
  millie_f_v_c %>% mutate(Label = "PlaqueFar_vs_Control"),
  millie_n_v_c %>% mutate(Label = "PlaqueNear_vs_Control"),
  millie_n_v_f %>% mutate(Label = "PlaqueNear_vs_PlaqueFar")
) %>%
  rename(
    log2FC_limma = logFC,
    adj.pvalue_limma = adj.P.Val
  ) %>%
  mutate(
    Status_limma = case_when(
      adj.pvalue_limma < 0.05 & log2FC_limma > 0.5 ~ "Upregulated",
      adj.pvalue_limma < 0.05 & log2FC_limma < -0.5 ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  ) %>%
  select(Gene, Label, log2FC_limma, adj.pvalue_limma, Status_limma)

millie_combined

write.csv(millie_combined, file.path(millie_dir, "millie_combined.csv"))
```

## combine both msstats and limma dataframes

```{r}
comparison_df <- msstats_ready %>%
  select(Protein, Gene, Label, log2FC_msstats = log2FC, adj.pvalue_msstats = adj.pvalue, Status_msstats = Status) %>%
  full_join(millie_combined, by = c("Gene", "Label")) %>%
  mutate(
    # Create a column to easily find discrepancies
    Agreement = case_when(
      Status_msstats == Status_limma ~ "Consistent",
      Status_msstats == "Not Significant" | Status_limma == "Not Significant" ~ "Pipeline Specific",
      TRUE ~ "Contradictory" # One says Up, one says Down
    )
  )

# Preview the overlap
table(comparison_df$Status_msstats, comparison_df$Status_limma)

comparison_df %>% filter(Gene== "Cd44")
```

## Plot concordance results

```{r}

plot_data <- comparison_df %>%
  # 1. Determine Significance categories based on your previous Status columns
  mutate(
    Significance = case_when(
      Status_msstats != "Not Significant" & Status_limma != "Not Significant" ~ "Significant in Both",
      Status_msstats != "Not Significant" & Status_limma == "Not Significant" ~ "Significant in MSstats only",
      Status_msstats == "Not Significant" & Status_limma != "Not Significant" ~ "Significant in limma only",
      TRUE ~ "Not Significant"
    ),
    # 2. Calculate distance from origin (Magnitude in both) to pick top labels
    # Using coalesce to handle NAs just in case a protein is only in one pipeline
    dist = sqrt(coalesce(log2FC_msstats, 0)^2 + coalesce(log2FC_limma, 0)^2)
  )

# Define your colors to match the exact spelling of categories
sig_colors <- c(
  "Significant in Both"         = "purple",
  "Significant in MSstats only" = "#00BFC4", # Cyan/Blue
  "Significant in limma only"   = "#F8766D", # Coral/Red
  "Not Significant"             = "grey80"
)

# --- STEP 2: Loop through each comparison and plot ---
comparisons <- unique(plot_data$Label)

for (comp in comparisons) {
  
  cat("Plotting:", comp, "\n")
  
  # Subset data for the current comparison
  df_sub <- plot_data %>% filter(Label == comp)
  
  # Check if we have data to plot before proceeding
  if(nrow(df_sub %>% filter(!is.na(log2FC_msstats) & !is.na(log2FC_limma))) == 0) {
    message(paste("No overlapping data found for:", comp, "- skipping."))
    next 
  }
  
  # Get top 30 genes to label based on distance from origin
  genes_to_label <- df_sub %>%
    filter(Significance != "Not Significant") %>%
    slice_max(order_by = dist, n = 30, with_ties = FALSE)
  
  # Calculate Pearson correlation for the subtitle 
  cor_val <- round(cor(df_sub$log2FC_msstats, df_sub$log2FC_limma, use = "complete.obs"), 3)
  
  # Clean up the comparison name for the plot title (e.g., "PlaqueNear_vs_Control" -> "Plaque Near vs Control")
  clean_title <- gsub("_", " ", comp)
  # Optional: Add spaces before capital letters if your label is CamelCase
  # clean_title <- gsub("([a-z])([A-Z])", "\\1 \\2", clean_title) 
  
  
  #### settings to make plot look pretty -> so all genes fit nicely in plot
  x_bulk <- quantile(df_sub$log2FC_limma, probs = c(0.01, 0.99), na.rm = TRUE)
  y_bulk <- quantile(df_sub$log2FC_msstats, probs = c(0.01, 0.99), na.rm = TRUE)
  
  #Get the range of the proteins we actually want to label
  x_label_range <- range(genes_to_label$log2FC_limma, na.rm = TRUE)
  y_label_range <- range(genes_to_label$log2FC_msstats, na.rm = TRUE)
  
  # 3. Final Limits: Use the wider of the two ranges so labels aren't cut off
  final_xlim <- c(min(x_bulk[1], x_label_range[1]), max(x_bulk[2], x_label_range[2]))
  final_ylim <- c(min(y_bulk[1], y_label_range[1]), max(y_bulk[2], y_label_range[2]))
  
  # Add 15% padding so labels have room to "repel" into
  x_pad <- diff(final_xlim) * 0.15
  y_pad <- diff(final_ylim) * 0.15
  
  
  
  # Build The Plot
  p <- ggplot(df_sub, aes(x = log2FC_limma, y = log2FC_msstats)) +
    
    # 1. Diagonal and Axes
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "solid", linewidth = 0.5) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    
    # 2. Scatter Points
    geom_point(aes(color = Significance), alpha = 0.6, size = 2) +
    
    # 3. Optimized Labels (Mapped to your "Gene" column)
    geom_label_repel(
      data = genes_to_label, 
      aes(label = Gene, color = Significance),
      size = 4,                  
      fontface = "bold",         
      box.padding = 0.5,         
      point.padding = 0.3,       
      force = 2,                 
      max.overlaps = Inf,        
      min.segment.length = 0,    
      show.legend = FALSE,
      fill = "white"
    ) +
    
    # 4. Styling
    scale_color_manual(values = sig_colors) +
    coord_cartesian(
      xlim = c(final_xlim[1] - x_pad, final_xlim[2] + x_pad),
      ylim = c(final_ylim[1] - y_pad, final_ylim[2] + y_pad),
      expand = FALSE # Keeps our manual padding exact
    ) + 
    theme_bw(base_size = 16) + 
    
    labs(
      title = paste("Concordance:", clean_title),
      subtitle = paste("limma vs. MSstats | Pearson r =", cor_val),
      x = "limma Log2FC",
      y = "MSstats Log2FC",
      color = "Significance (FDR < 0.05)"
    ) +
    
    theme(
      legend.position = "bottom",
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 18),
      axis.title = element_text(face = "bold")
    ) + 
    guides(color = guide_legend(
      override.aes = list(size = 5), 
      nrow = 2
    ))
  
  # Print to RStudio viewer/console so you can see them as they generate
  print(p)
  
  # Sanitize filename (replace spaces/special chars with underscores)
  safe_filename <- str_replace_all(comp, "[^A-Za-z0-9]+", "_")
  safe_filename <- str_replace(safe_filename, "_$", "")
  
  # Save to PDF
  ggsave(
    filename = file.path(graphs_dir, paste0("MSstats_vs_limma_concordance_", safe_filename, ".pdf")),
    plot = p,
    width = 10,
    height = 10,
    dpi = 300
  )
}

cat("All plots generated and saved successfully to:", graphs_dir, "\n")
```

## Print summary table comparing the two methods

Summarises the overlap in detection and significance.

```{r}
summary_table <- comparison_df %>%
  group_by(Label) %>%
  summarise(
    # 1. Total Detected overall (union of both methods)
    Total_Detected_Overall = n(),
    
    # 2. Detected by specific methods
    Detected_MSstats = sum(!is.na(log2FC_msstats)),
    Detected_limma   = sum(!is.na(log2FC_limma)),
    
    # 3. Both Detected (Shared column)
    Both_Detected    = sum(!is.na(log2FC_msstats) & !is.na(log2FC_limma)),
    
    # 4. Significant (FDR < 0.05)
    Sig_MSstats      = sum(Status_msstats != "Not Significant", na.rm = TRUE),
    Sig_limma        = sum(Status_limma != "Not Significant", na.rm = TRUE),
    
    # 5. Both Significant (Shared column)
    Both_Significant = sum(Status_msstats != "Not Significant" & 
                           Status_limma != "Not Significant", na.rm = TRUE)
  ) %>%
  # Clean up the Label for readability
  mutate(Label = gsub("_", " ", Label))

summary_table
```

## How many signficant proteins from MSstats pipeline did the limma pipeline detect as present?

```{r}
#Filter for signficant msstats proteins and check if detected by Millie
msstats_sig_analysis <- comparison_df %>%
  filter(Status_msstats != "Not Significant") %>%
  mutate(
    Detected_by_Millie = ifelse(!is.na(log2FC_limma), "Yes", "No")
  )

# Summarize the findings per comparison
detection_overlap <- msstats_sig_analysis %>%
  group_by(Label, Detected_by_Millie) %>%
  tally(name = "Protein_Count") %>%
  tidyr::pivot_wider(names_from = Detected_by_Millie, values_from = Protein_Count, names_prefix = "Detected_by_limma_")

print(detection_overlap)
```

### What were the signficant proteins that were detected in the limma analysis but not msstats?

These proteins all seem like very solid candidates.

Limma seems to be more powerful for the near vs far comparison

```{r}
limma_exclusive_sig <- comparison_df %>%
  filter(Label == "PlaqueNear_vs_PlaqueFar") %>%
  filter(Status_limma != "Not Significant" & Status_msstats == "Not Significant") %>%
  select(Gene, 
         log2FC_limma, adj.pvalue_limma, 
         log2FC_msstats, adj.pvalue_msstats) %>%
  arrange(adj.pvalue_limma)

limma_exclusive_sig
```

### What were the signficant proteins that were detected in the msstats analysis but not limma?

```{r}
msstats_exclusive_sig <- comparison_df %>%
  filter(Label == "PlaqueNear_vs_PlaqueFar") %>%
  filter(Status_limma == "Not Significant" & Status_msstats != "Not Significant") %>%
  select(Gene, 
         log2FC_limma, adj.pvalue_limma, 
         log2FC_msstats, adj.pvalue_msstats) %>%
  arrange(adj.pvalue_limma)

msstats_exclusive_sig
```


```{r}
msstats_exclusive_sig %>%
  pull(Gene)
```




### Check for how many MSstats proteins were FDR \<0.05 but \< 0.5 log2FC

```{r}
check_cutoffs_msstats <- comparison_df %>%
  filter(Label == "PlaqueNear_vs_PlaqueFar") %>%
  filter(adj.pvalue_msstats < 0.05 & Status_msstats == "Not Significant") %>%
  select(Gene, log2FC_msstats, adj.pvalue_msstats)

check_cutoffs_msstats
```

### Check for how many limma proteins were FDR \<0.05 but \< 0.5 log2FC

```{r}
check_cutoffs_limma <- comparison_df %>%
  filter(Label == "PlaqueNear_vs_PlaqueFar") %>%
  filter(adj.pvalue_msstats < 0.05 & Status_limma == "Not Significant") %>%
  select(Gene, log2FC_limma, adj.pvalue_limma, log2FC_msstats, adj.pvalue_msstats)

check_cutoffs_limma
```


```{r}
 
```
