---
title: "6b. Pipeline-Exclusive DE Proteins: Raw & Normalised Boxplots"
format: html
---

==================================================================

# Pipeline-Exclusive DE Protein Visualisation

Author: Carlo Zanetti

This script identifies proteins called as differentially expressed by one pipeline but **not** the other, and visualises their abundance distributions using both raw Spectronaut intensities and MSstats-normalised values. This serves as a sense-check for pipeline-specific hits and helps diagnose where and why the two methods diverge.

### Workflow overview

1.  **Rebuild comparison_df** merging MSstats and limma results
2.  **Extract pipeline-exclusive hits** in both directions (MSstats-only and limma-only)
3.  **Prepare MSstats normalised** ProteinLevelData abundances
4.  **Prepare raw Spectronaut** PG.Quantity values (long format, log2-transformed, batch-filtered)
5.  **Generate faceted boxplots** for both data sources, per comparison, per direction
6.  **Export summary tables** with side-by-side pipeline statistics

==================================================================

## Load libraries

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(qs2)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(stringr)
})
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run6"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
millie_dir  <- file.path(base_dir, "data", "processed_millie")
graphs_dir  <- file.path(results_dir, "concordance")

boxplot_dir <- file.path(graphs_dir, "pipeline_exclusive_boxplots")
dir.create(boxplot_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load required objects

```{r}
res_df         <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))
clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))
clean_meta     <- fread(file.path(base_dir, "data/processed/run1/clean_metadata.csv"))

millie_f_v_c <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_far_vs_Control.csv"))
millie_n_v_c <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_near_vs_Control.csv"))
millie_n_v_f <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_near_vs_Far.csv"))

protein_dictionary <- clean_spec_raw %>%
  dplyr::select(Protein = PG.ProteinGroups,
                Gene = PG.Genes,
                Description = PG.ProteinDescriptions) %>%
  distinct()
```

## Rebuild comparison_df

```{r}
msstats_ready <- res_df %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  left_join(protein_dictionary, by = "Protein") %>%
  mutate(
    Gene = ifelse(is.na(Gene) | Gene == "", Protein, Gene),
    Status = case_when(
      adj.pvalue < 0.05 & log2FC > 0.5  ~ "Upregulated",
      adj.pvalue < 0.05 & log2FC < -0.5 ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )

millie_combined <- bind_rows(
  millie_f_v_c %>% mutate(Label = "PlaqueFar_vs_Control"),
  millie_n_v_c %>% mutate(Label = "PlaqueNear_vs_Control"),
  millie_n_v_f %>% mutate(Label = "PlaqueNear_vs_PlaqueFar")
) %>%
  dplyr::rename(log2FC_limma = logFC, adj.pvalue_limma = adj.P.Val) %>%
  mutate(
    Status_limma = case_when(
      adj.pvalue_limma < 0.05 & log2FC_limma > 0.5  ~ "Upregulated",
      adj.pvalue_limma < 0.05 & log2FC_limma < -0.5 ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  ) %>%
  dplyr::select(Gene, Label, log2FC_limma, adj.pvalue_limma, Status_limma)

comparison_df <- msstats_ready %>%
  dplyr::select(Protein, Gene, Label,
                log2FC_msstats = log2FC,
                adj.pvalue_msstats = adj.pvalue,
                Status_msstats = Status) %>%
  full_join(millie_combined, by = c("Gene", "Label"))
```

## Extract pipeline-exclusive significant proteins

### MSstats-exclusive: significant in MSstats, not in limma

```{r}
msstats_exclusive <- comparison_df %>%
  filter(
    Status_msstats != "Not Significant",
    Status_limma == "Not Significant" | is.na(Status_limma)
  ) %>%
  group_by(Label) %>%
  arrange(adj.pvalue_msstats, .by_group = TRUE) %>%
  slice_head(n = 20) %>%
  ungroup() %>%
  dplyr::select(
    Protein, Gene, Label,
    log2FC_msstats, adj.pvalue_msstats, Status_msstats,
    log2FC_limma, adj.pvalue_limma, Status_limma
  )

cat("MSstats-exclusive top 20 per comparison:\n")
as.data.frame(msstats_exclusive %>% dplyr::count(Label))
```

### Limma-exclusive: significant in limma, not in MSstats

For these proteins we need to recover the Protein ID from the dictionary since limma results only carry Gene names. Where a limma-exclusive gene has no matching Protein ID (i.e. it wasn't in the MSstats results at all), we match back via the protein dictionary.

```{r}
limma_exclusive <- comparison_df %>%
  filter(
    Status_limma != "Not Significant",
    Status_msstats == "Not Significant" | is.na(Status_msstats)
  ) %>%
  # For proteins only in limma, Protein column may be NA — recover from dictionary
  left_join(
    protein_dictionary %>% dplyr::select(Protein, Gene_dict = Gene),
    by = "Protein"
  ) %>%
  mutate(
    Protein = case_when(
      !is.na(Protein) ~ Protein,
      TRUE ~ {
        matched <- protein_dictionary$Protein[match(Gene, protein_dictionary$Gene)]
        matched
      }
    )
  ) %>%
  dplyr::select(-Gene_dict) %>%
  group_by(Label) %>%
  arrange(adj.pvalue_limma, .by_group = TRUE) %>%
  slice_head(n = 20) %>%
  ungroup() %>%
  dplyr::select(
    Protein, Gene, Label,
    log2FC_msstats, adj.pvalue_msstats, Status_msstats,
    log2FC_limma, adj.pvalue_limma, Status_limma
  )

cat("\nLimma-exclusive top 20 per comparison:\n")
as.data.frame(limma_exclusive %>% dplyr::count(Label))
```

```{r}
fwrite(msstats_exclusive, file.path(boxplot_dir, "msstats_exclusive_top20.csv"))
fwrite(limma_exclusive, file.path(boxplot_dir, "limma_exclusive_top20.csv"))
```

## Prepare MSstats normalised protein-level data

Loaded first so we can verify batch/run alignment with the raw data.

```{r}
processed_data <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))

msstats_abundance <- processed_data$ProteinLevelData %>%
  as.data.frame() %>%
  dplyr::rename(
    Condition      = GROUP,
    BioReplicate   = SUBJECT,
    Run            = RUN,
    LogIntensities = LogIntensities
  )

cat("MSstats abundance rows:", nrow(msstats_abundance), "\n")
cat("MSstats unique runs:", n_distinct(msstats_abundance$Run), "\n")
```

## Prepare raw Spectronaut protein-level data (long format)

We filter to batches 1, 3, 4 via metadata to match the samples used by MSstats. We cannot match on Run names directly because MSstats strips the `[nnn] ` prefix and `.d.PG.Quantity` suffix during processing.

```{r}
anno_cols    <- c("PG.ProteinGroups", "PG.Genes", "PG.ProteinDescriptions")
run_cols_raw <- setdiff(colnames(clean_spec_raw), anno_cols)

# Batches retained after QC in the MSstats workflow
good_batches <- c(1, 3, 4)

spec_long <- clean_spec_raw %>%
  tidyr::pivot_longer(
    cols      = all_of(run_cols_raw),
    names_to  = "Run",
    values_to = "Intensity"
  ) %>%
  dplyr::rename(Protein = PG.ProteinGroups) %>%
  left_join(
    clean_meta %>% dplyr::select(Run, Condition, BioReplicate, Batch),
    by = "Run"
  ) %>%
  filter(
    !is.na(Condition),
    Batch %in% good_batches
  ) %>%
  mutate(
    Log2_Intensity = ifelse(!is.na(Intensity) & Intensity > 0, log2(Intensity), NA_real_)
  )

cat("Spectronaut long format rows:", nrow(spec_long), "\n")
cat("Unique runs after batch filtering:", n_distinct(spec_long$Run), "\n")
cat("Runs per condition:\n")
as.data.frame(spec_long %>% distinct(Run, Condition) %>% dplyr::count(Condition))
```

### Sanity check: do exclusive protein IDs exist in spec_long?

```{r}
msstats_excl_proteins <- unique(msstats_exclusive$Protein)
limma_excl_proteins   <- unique(limma_exclusive$Protein)
spec_proteins         <- unique(spec_long$Protein)

cat("MSstats-exclusive proteins found in spec_long:",
    sum(msstats_excl_proteins %in% spec_proteins), "/",
    length(msstats_excl_proteins), "\n")
cat("Limma-exclusive proteins found in spec_long:",
    sum(limma_excl_proteins %in% spec_proteins), "/",
    length(limma_excl_proteins), "\n")

# If any are missing, show which ones
missing_msstats <- msstats_excl_proteins[!msstats_excl_proteins %in% spec_proteins]
missing_limma   <- limma_excl_proteins[!limma_excl_proteins %in% spec_proteins]

if (length(missing_msstats) > 0) {
  cat("\nMissing MSstats-exclusive proteins:\n")
  print(missing_msstats)
}
if (length(missing_limma) > 0) {
  cat("\nMissing Limma-exclusive proteins:\n")
  print(missing_limma)
}
```

## Define plotting function

Generalised to handle both directions. The `pipeline_label` argument controls titles and file naming.

```{r}
cond_colors <- c(
  "control"     = "#1b9e77",
  "plaque_near" = "#d95f02",
  "plaque_far"  = "#7570b3"
)

plot_exclusive_boxplots <- function(proteins_df, spec_long_df, msstats_abund_df,
                                    comparison_label, output_dir,
                                    pipeline_label = "MSstats-exclusive",
                                    rank_by_col = "adj.pvalue_msstats") {

  genes_to_plot <- proteins_df %>%
    filter(Label == comparison_label) %>%
    arrange(.data[[rank_by_col]])

  if (nrow(genes_to_plot) == 0) {
    message("No ", pipeline_label, " proteins for: ", comparison_label)
    return(invisible(NULL))
  }

  cat("\n--- Plotting", nrow(genes_to_plot), pipeline_label,
      "proteins for:", comparison_label, "---\n")

  safe_comp   <- str_replace_all(comparison_label, "[^A-Za-z0-9]+", "_")
  safe_label  <- str_replace_all(pipeline_label, "[^A-Za-z0-9]+", "_")
  clean_title <- gsub("_", " ", comparison_label)

  # Build facet labels showing stats from BOTH pipelines
  make_facet_label <- function(df) {
    df %>%
      mutate(
        msstats_info = ifelse(
          !is.na(adj.pvalue_msstats),
          paste0("MS: FC=", round(log2FC_msstats, 2),
                 " p=", formatC(adj.pvalue_msstats, format = "e", digits = 1)),
          "MS: not tested"
        ),
        limma_info = ifelse(
          !is.na(adj.pvalue_limma),
          paste0("LM: FC=", round(log2FC_limma, 2),
                 " p=", formatC(adj.pvalue_limma, format = "e", digits = 1)),
          "LM: not tested"
        ),
        facet_label = paste0(Gene, "\n", msstats_info, "\n", limma_info)
      )
  }

  order_facets <- function(plot_df, genes_df, rank_col) {
    ordered_genes <- genes_df %>% arrange(.data[[rank_col]]) %>% pull(Gene)
    facet_map <- plot_df %>%
      dplyr::select(Gene, facet_label) %>% distinct()
    ordered_labels <- left_join(
      data.frame(Gene = ordered_genes), facet_map, by = "Gene"
    ) %>% pull(facet_label) %>% na.omit() %>% unique()
    plot_df$facet_label <- factor(plot_df$facet_label, levels = ordered_labels)
    plot_df
  }

  # ---- A) Raw Spectronaut boxplots ----
  raw_plot_data <- spec_long_df %>%
    filter(Protein %in% genes_to_plot$Protein) %>%
    left_join(
      genes_to_plot %>% dplyr::select(Protein, Gene,
                                       log2FC_msstats, adj.pvalue_msstats,
                                       log2FC_limma, adj.pvalue_limma),
      by = "Protein"
    ) %>%
    filter(!is.na(Log2_Intensity))

  cat("  Raw plot data rows:", nrow(raw_plot_data), "\n")

  if (nrow(raw_plot_data) > 0) {
    raw_plot_data <- raw_plot_data %>%
      make_facet_label() %>%
      order_facets(genes_to_plot, rank_by_col)

    n_facets <- n_distinct(raw_plot_data$facet_label)

    p_raw <- ggplot(raw_plot_data,
                    aes(x = Condition, y = Log2_Intensity, fill = Condition)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.6) +
      geom_jitter(width = 0.15, size = 1.5, alpha = 0.7) +
      facet_wrap(~ facet_label, scales = "free_y", ncol = 5) +
      scale_fill_manual(values = cond_colors) +
      theme_bw(base_size = 12) +
      labs(
        title    = paste0("Raw Spectronaut: ", pipeline_label, " DE Proteins"),
        subtitle = paste0(clean_title, " | Top ", nrow(genes_to_plot), " by FDR"),
        x = NULL,
        y = "Log2(PG.Quantity)"
      ) +
      theme(
        axis.text.x    = element_text(angle = 45, hjust = 1, size = 8),
        strip.text     = element_text(size = 6, face = "bold"),
        legend.position = "bottom",
        plot.title     = element_text(face = "bold", size = 14),
        plot.subtitle  = element_text(size = 10, color = "grey40")
      )

    print(p_raw)

    ggsave(
      file.path(output_dir, paste0("Raw_", safe_label, "_", safe_comp, ".pdf")),
      plot = p_raw,
      width = 16, height = max(ceiling(n_facets / 5) * 4, 5),
      dpi = 300, limitsize = FALSE
    )
  } else {
    cat("  WARNING: No raw data found for these proteins. Check Protein ID format.\n")
  }

  # ---- B) MSstats normalised boxplots ----
  norm_plot_data <- msstats_abund_df %>%
    filter(Protein %in% genes_to_plot$Protein) %>%
    left_join(
      genes_to_plot %>% dplyr::select(Protein, Gene,
                                       log2FC_msstats, adj.pvalue_msstats,
                                       log2FC_limma, adj.pvalue_limma),
      by = "Protein"
    ) %>%
    filter(!is.na(LogIntensities))

  cat("  Normalised plot data rows:", nrow(norm_plot_data), "\n")

  if (nrow(norm_plot_data) > 0) {
    norm_plot_data <- norm_plot_data %>%
      make_facet_label() %>%
      order_facets(genes_to_plot, rank_by_col)

    n_facets_norm <- n_distinct(norm_plot_data$facet_label)

    p_norm <- ggplot(norm_plot_data,
                     aes(x = Condition, y = LogIntensities, fill = Condition)) +
      geom_boxplot(outlier.shape = NA, alpha = 0.7, width = 0.6) +
      geom_jitter(width = 0.15, size = 1.5, alpha = 0.7) +
      facet_wrap(~ facet_label, scales = "free_y", ncol = 5) +
      scale_fill_manual(values = cond_colors) +
      theme_bw(base_size = 12) +
      labs(
        title    = paste0("MSstats Normalised: ", pipeline_label, " DE Proteins"),
        subtitle = paste0(clean_title, " | Top ", nrow(genes_to_plot), " by FDR"),
        x = NULL,
        y = "Log2 Intensity (TMP normalised)"
      ) +
      theme(
        axis.text.x    = element_text(angle = 45, hjust = 1, size = 8),
        strip.text     = element_text(size = 6, face = "bold"),
        legend.position = "bottom",
        plot.title     = element_text(face = "bold", size = 14),
        plot.subtitle  = element_text(size = 10, color = "grey40")
      )

    print(p_norm)

    ggsave(
      file.path(output_dir, paste0("Normalised_", safe_label, "_", safe_comp, ".pdf")),
      plot = p_norm,
      width = 16, height = max(ceiling(n_facets_norm / 5) * 4, 5),
      dpi = 300, limitsize = FALSE
    )
  } else {
    cat("  WARNING: No normalised data found for these proteins. Check Protein ID format.\n")
  }
}
```

## Generate boxplots: MSstats-exclusive proteins

```{r, fig.width=16, fig.height=14}
for (comp in unique(msstats_exclusive$Label)) {
  plot_exclusive_boxplots(
    proteins_df      = msstats_exclusive,
    spec_long_df     = spec_long,
    msstats_abund_df = msstats_abundance,
    comparison_label = comp,
    output_dir       = boxplot_dir,
    pipeline_label   = "MSstats-exclusive",
    rank_by_col      = "adj.pvalue_msstats"
  )
}
```

## Generate boxplots: Limma-exclusive proteins

```{r, fig.width=16, fig.height=14}
for (comp in unique(limma_exclusive$Label)) {
  plot_exclusive_boxplots(
    proteins_df      = limma_exclusive,
    spec_long_df     = spec_long,
    msstats_abund_df = msstats_abundance,
    comparison_label = comp,
    output_dir       = boxplot_dir,
    pipeline_label   = "Limma-exclusive",
    rank_by_col      = "adj.pvalue_limma"
  )
}
```

## Summary tables

```{r}
make_summary <- function(exclusive_df, calling_pipeline = "msstats") {

  other_pipeline <- ifelse(calling_pipeline == "msstats", "limma", "msstats")
  calling_fdr    <- paste0("adj.pvalue_", calling_pipeline)
  other_fdr      <- paste0("adj.pvalue_", other_pipeline)
  calling_fc     <- paste0("log2FC_", calling_pipeline)
  other_fc       <- paste0("log2FC_", other_pipeline)

  exclusive_df %>%
    mutate(
      direction = ifelse(.data[[calling_fc]] > 0, "Up", "Down"),
      other_detected = ifelse(!is.na(.data[[other_fdr]]), "Yes", "No"),
      other_note = case_when(
        is.na(.data[[other_fdr]])   ~ "Not detected",
        .data[[other_fdr]] < 0.1    ~ paste0("Trend (FDR=",
                                              formatC(.data[[other_fdr]], format = "e", digits = 1), ")"),
        TRUE ~ paste0("NS (FDR=", round(.data[[other_fdr]], 3), ")")
      )
    ) %>%
    dplyr::select(
      Label, Gene, Protein, direction,
      all_of(c(calling_fc, calling_fdr, other_fc, other_fdr)),
      other_note
    ) %>%
    arrange(Label, .data[[calling_fdr]])
}
```

### MSstats-exclusive summary

```{r}
summary_msstats_excl <- make_summary(msstats_exclusive, "msstats")
fwrite(summary_msstats_excl, file.path(boxplot_dir, "summary_msstats_exclusive.csv"))
summary_msstats_excl
```

### Limma-exclusive summary

```{r}
summary_limma_excl <- make_summary(limma_exclusive, "limma")
fwrite(summary_limma_excl, file.path(boxplot_dir, "summary_limma_exclusive.csv"))
summary_limma_excl
```

### Quick diagnostic: how many limma-exclusive hits show a trend in MSstats?

This helps distinguish proteins where MSstats genuinely disagrees from borderline cases where MSstats was close to calling significance.

```{r}
limma_exclusive %>%
  mutate(
    msstats_category = case_when(
      is.na(adj.pvalue_msstats)      ~ "Not in MSstats",
      adj.pvalue_msstats < 0.1       ~ "MSstats trend (FDR<0.1)",
      adj.pvalue_msstats < 0.2       ~ "MSstats marginal (FDR 0.1-0.2)",
      TRUE                           ~ "MSstats NS (FDR>0.2)"
    )
  ) %>%
  dplyr::count(Label, msstats_category) %>%
  tidyr::pivot_wider(names_from = msstats_category, values_from = n, values_fill = 0) %>%
  as.data.frame()
```

```{r}
cat("\nAll pipeline-exclusive boxplots saved to:", boxplot_dir, "\n")
```