---
title: "6c. Variance Diagnostics: MSstats vs limma"
format: html
---

==================================================================

# Variance Diagnostic: Why does MSstats lose sensitivity for Near vs Far?

Author: Carlo Zanetti

This script tests the hypothesis that MSstats' per-protein variance estimates are inflated for the Plaque Near vs Far comparison (where both groups are high-variance) relative to limma's empirical Bayes moderated estimates, explaining why MSstats is more conservative for that specific comparison.

### Predictions

1. MSstats standard errors should be larger for Near vs Far than for the vs Control comparisons
2. Limma's moderated variance should be more stable across comparisons
3. The proteins lost by MSstats in Near vs Far should be those with the highest MSstats variance inflation
4. Log2FC agreement between pipelines should remain high — it's the variance, not the effect size, that diverges

==================================================================

## Load libraries

```{r}
suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(stringr)
  library(qs2)
})
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run5"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
millie_dir  <- file.path(base_dir, "data", "processed_millie")
graphs_dir  <- file.path(results_dir, "concordance")

variance_dir <- file.path(graphs_dir, "variance_diagnostics")
dir.create(variance_dir, recursive = TRUE, showWarnings = FALSE)
```

## Load data

```{r}
# MSstats full results (contains SE column)
res_df <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))

# Limma results
millie_f_v_c <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_far_vs_Control.csv"))
millie_n_v_c <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_near_vs_Control.csv"))
millie_n_v_f <- read.csv(file.path(millie_dir, "limma_all_proteins_Plaque_near_vs_Far.csv"))

# Protein dictionary
clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))
protein_dictionary <- clean_spec_raw %>%
  dplyr::select(Protein = PG.ProteinGroups,
                Gene = PG.Genes) %>%
  distinct()
```

## Prepare MSstats data

The `SE` column in the MSstats output is the standard error of the log2FC estimate — this directly reflects the per-protein variance.

```{r}
msstats_data <- res_df %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  left_join(protein_dictionary, by = "Protein") %>%
  mutate(Gene = ifelse(is.na(Gene) | Gene == "", Protein, Gene)) %>%
  dplyr::select(Gene, Protein, Label,
                log2FC_msstats = log2FC,
                SE_msstats = SE,
                pvalue_msstats = pvalue,
                adj.pvalue_msstats = adj.pvalue,
                DF_msstats = DF)
```

## Prepare limma data

Limma's `t` statistic and `logFC` let us back-calculate the effective SE: SE = logFC / t

```{r}
limma_combined <- bind_rows(
  millie_f_v_c %>% mutate(Label = "PlaqueFar_vs_Control"),
  millie_n_v_c %>% mutate(Label = "PlaqueNear_vs_Control"),
  millie_n_v_f %>% mutate(Label = "PlaqueNear_vs_PlaqueFar")
) %>%
  mutate(
    SE_limma = logFC / t
  ) %>%
  dplyr::select(Gene, Label,
                log2FC_limma = logFC,
                SE_limma,
                t_limma = t,
                pvalue_limma = P.Value,
                adj.pvalue_limma = adj.P.Val)
```

## Merge pipelines

```{r}
merged <- msstats_data %>%
  inner_join(limma_combined, by = c("Gene", "Label"))

cat("Merged proteins per comparison:\n")
as.data.frame(merged %>% dplyr::count(Label))
```

## Test 1: Are MSstats SEs larger for Near vs Far?

If our hypothesis is correct, MSstats SE should be systematically higher for PlaqueNear_vs_PlaqueFar compared to the vs Control comparisons, while limma's SE should be more stable.

```{r}
se_summary <- merged %>%
  group_by(Label) %>%
  summarise(
    median_SE_msstats = median(SE_msstats, na.rm = TRUE),
    mean_SE_msstats   = mean(SE_msstats, na.rm = TRUE),
    median_SE_limma   = median(SE_limma, na.rm = TRUE),
    mean_SE_limma     = mean(SE_limma, na.rm = TRUE),
    # Ratio shows how much bigger MSstats SE is relative to limma
    median_SE_ratio   = median(SE_msstats / abs(SE_limma), na.rm = TRUE),
    .groups = "drop"
  )

as.data.frame(se_summary)
```

## Test 2: Distribution of SE ratio across comparisons

The SE ratio (MSstats SE / limma SE) should be higher for Near vs Far if MSstats variance is specifically inflated for that comparison.

```{r, fig.width=10, fig.height=5}
merged <- merged %>%
  mutate(SE_ratio = SE_msstats / abs(SE_limma))

p_se_ratio <- ggplot(merged, aes(x = Label, y = SE_ratio, fill = Label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.2) +
  coord_cartesian(ylim = c(0, quantile(merged$SE_ratio, 0.95, na.rm = TRUE))) +
  theme_bw(base_size = 14) +
  labs(
    title = "MSstats / limma Standard Error Ratio by Comparison",
    subtitle = "Higher ratio = MSstats more uncertain relative to limma",
    x = NULL,
    y = "SE ratio (MSstats / limma)"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))

print(p_se_ratio)

ggsave(file.path(variance_dir, "SE_ratio_by_comparison.pdf"),
       plot = p_se_ratio, width = 10, height = 5)
```

### Statistical test: is the SE ratio significantly higher for Near vs Far?

```{r}
nf_ratios <- merged %>% filter(Label == "PlaqueNear_vs_PlaqueFar") %>% pull(SE_ratio)
nc_ratios <- merged %>% filter(Label == "PlaqueNear_vs_Control") %>% pull(SE_ratio)
fc_ratios <- merged %>% filter(Label == "PlaqueFar_vs_Control") %>% pull(SE_ratio)

cat("Near vs Far  median SE ratio:", round(median(nf_ratios, na.rm = TRUE), 3), "\n")
cat("Near vs Ctrl median SE ratio:", round(median(nc_ratios, na.rm = TRUE), 3), "\n")
cat("Far vs Ctrl  median SE ratio:", round(median(fc_ratios, na.rm = TRUE), 3), "\n\n")

# Wilcoxon tests
cat("Near vs Far vs Near vs Control:\n")
print(wilcox.test(nf_ratios, nc_ratios, alternative = "greater"))

cat("\nNear vs Far vs Far vs Control:\n")
print(wilcox.test(nf_ratios, fc_ratios, alternative = "greater"))
```

## Test 3: SE scatter plots — MSstats vs limma per comparison

Points should fall closer to the diagonal for vs Control comparisons and above it (MSstats > limma) for Near vs Far.

```{r, fig.width=14, fig.height=5}
p_se_scatter <- ggplot(merged, aes(x = abs(SE_limma), y = SE_msstats)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ Label, scales = "free") +
  theme_bw(base_size = 12) +
  labs(
    title = "Standard Error Comparison: MSstats vs limma",
    subtitle = "Points above the red line = MSstats more uncertain",
    x = "limma SE (absolute)",
    y = "MSstats SE"
  ) +
  theme(strip.text = element_text(face = "bold"))

print(p_se_scatter)

ggsave(file.path(variance_dir, "SE_scatter_by_comparison.pdf"),
       plot = p_se_scatter, width = 14, height = 5)
```

## Test 4: Do lost proteins have inflated MSstats variance?

For the Near vs Far comparison, compare SE distributions for proteins that are significant in both pipelines vs limma-exclusive (significant in limma, lost by MSstats).

```{r, fig.width=10, fig.height=5}
nf_data <- merged %>%
  filter(Label == "PlaqueNear_vs_PlaqueFar") %>%
  mutate(
    sig_msstats = adj.pvalue_msstats < 0.05 & abs(log2FC_msstats) > 0.5,
    sig_limma   = adj.pvalue_limma < 0.05 & abs(log2FC_limma) > 0.5,
    category = case_when(
      sig_msstats & sig_limma  ~ "Significant in both",
      !sig_msstats & sig_limma ~ "Limma-exclusive",
      sig_msstats & !sig_limma ~ "MSstats-exclusive",
      TRUE                     ~ "Not significant"
    )
  )

# Focus on the informative comparison
plot_cats <- c("Significant in both", "Limma-exclusive")

p_lost <- ggplot(nf_data %>% filter(category %in% plot_cats),
                 aes(x = category, y = SE_msstats, fill = category)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.5) +
  theme_bw(base_size = 14) +
  scale_fill_manual(values = c("Significant in both" = "purple",
                                "Limma-exclusive" = "#F8766D")) +
  labs(
    title = "MSstats SE for Proteins Lost in Near vs Far",
    subtitle = "If hypothesis is correct, limma-exclusive proteins should have higher MSstats SE",
    x = NULL,
    y = "MSstats Standard Error"
  ) +
  theme(legend.position = "none")

print(p_lost)

ggsave(file.path(variance_dir, "SE_lost_proteins_near_vs_far.pdf"),
       plot = p_lost, width = 8, height = 5)

# Formal test
cat("Wilcoxon test: limma-exclusive vs both-significant MSstats SE\n")
both_se  <- nf_data %>% filter(category == "Significant in both") %>% pull(SE_msstats)
lost_se  <- nf_data %>% filter(category == "Limma-exclusive") %>% pull(SE_msstats)
print(wilcox.test(lost_se, both_se, alternative = "greater"))
```

## Test 5: Confirm fold changes agree — it's variance, not signal

If the issue is purely variance estimation, the log2FC values should still correlate well between pipelines even for Near vs Far.

```{r, fig.width=14, fig.height=5}
cors <- merged %>%
  group_by(Label) %>%
  summarise(
    pearson_r = round(cor(log2FC_msstats, log2FC_limma, use = "complete.obs"), 3),
    .groups = "drop"
  )

p_fc <- ggplot(merged, aes(x = log2FC_limma, y = log2FC_msstats)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  facet_wrap(~ Label) +
  geom_text(data = cors, aes(label = paste0("r = ", pearson_r)),
            x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
            size = 5, fontface = "bold", inherit.aes = FALSE) +
  theme_bw(base_size = 12) +
  labs(
    title = "Log2FC Agreement Between Pipelines",
    subtitle = "High correlation confirms the issue is variance estimation, not signal detection",
    x = "limma log2FC",
    y = "MSstats log2FC"
  ) +
  theme(strip.text = element_text(face = "bold"))

print(p_fc)

ggsave(file.path(variance_dir, "FC_correlation_by_comparison.pdf"),
       plot = p_fc, width = 14, height = 5)

as.data.frame(cors)
```

## Test 6: Degrees of freedom — are Near vs Far models underpowered?

Lower DF in MSstats means fewer effective observations, which inflates the t-distribution tails and makes it harder to reach significance.

```{r, fig.width=10, fig.height=5}
p_df <- ggplot(merged, aes(x = Label, y = DF_msstats, fill = Label)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 0.3, alpha = 0.2) +
  theme_bw(base_size = 14) +
  labs(
    title = "MSstats Degrees of Freedom by Comparison",
    subtitle = "Lower DF = wider t-distribution = harder to reach significance",
    x = NULL,
    y = "Satterthwaite DF"
  ) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))

print(p_df)

ggsave(file.path(variance_dir, "DF_by_comparison.pdf"),
       plot = p_df, width = 10, height = 5)

# Summary
merged %>%
  group_by(Label) %>%
  summarise(
    median_DF = median(DF_msstats, na.rm = TRUE),
    mean_DF   = round(mean(DF_msstats, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  as.data.frame()
```

## Summary

```{r}
cat("=== Variance Diagnostic Summary ===\n\n")
cat("If the hypothesis holds, you should see:\n")
cat("1. SE ratio (MSstats/limma) higher for Near vs Far than vs Control comparisons\n")
cat("2. SE scatter: points above diagonal specifically for Near vs Far\n")
cat("3. Limma-exclusive proteins have higher MSstats SE than both-significant proteins\n")
cat("4. Log2FC correlations remain high across all comparisons\n")
cat("5. DF may be lower for Near vs Far if more features are missing\n\n")
cat("All plots saved to:", variance_dir, "\n")
```