---
title: "07_cell_cell_interactions"
format: html
---

```{r}
library(OmnipathR)
library(dplyr)
library(tidyr)
library(data.table)
library(qs2)
```

## Set directories
```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Retrieve intercellullar communication database
```{r}
lr_network <- intercell_network(
  transmitter_param = list(categories = c('ligand')),
  receiver_param = list(categories = c('receptor')),
  resources = NULL,
  references_by_resource = TRUE)

lr_with_evidence <- lr_network %>%
    select(
        Ligand = source_genesymbol, 
        Receptor = target_genesymbol, 
        sources, 
        references
    ) %>%
    distinct() %>%
    # Count how many databases confirm this interaction
    mutate(curation_count = nchar(as.character(sources)) - nchar(gsub(";", "", as.character(sources))) + 1) %>%
    # Count how many individual PMIDs (papers) support it
    mutate(paper_count = nchar(as.character(references)) - nchar(gsub(";", "", as.character(references))) + 1) %>%
    # Replace NAs with 0
    mutate(across(c(curation_count, paper_count), ~replace_na(.x, 0)))

lr_with_evidence
# ) %>% 
#   # Keep only the gene symbols for the ligand (source) and receptor (target)
#   select(source_genesymbol, target_genesymbol) %>%
#   distinct() %>%
#   rename(Ligand = source_genesymbol, Receptor = target_genesymbol)
```
## Reaad in msstats results 
```{r}
res_df <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))

clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))

protein_dictionary <- clean_spec_raw %>%
  select(Protein = PG.ProteinGroups, 
         Gene = PG.Genes, 
         Description = PG.ProteinDescriptions) %>%
  distinct()

msstats_results <- res_df %>%
  # handle NAs
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  
  # Join with dictionary
  left_join(protein_dictionary, by = "Protein") %>%
  
  # Clean up Gene names and define significance
  mutate(
    # Fallback to Protein ID if Gene name is missing or empty
    #Change to upper case to match with human gene symbols
    Gene = toupper(ifelse(is.na(Gene) | Gene == "", Protein, Gene)),
  )
```

## Create ligand and receptor dataframes
```{r}
#Ligands data
ligands_data <- msstats_results %>%
  select(Label, Ligand = Gene, logFC_Ligand = log2FC, pval_Ligand = adj.pvalue)

#Receptor data
receptors_data <- msstats_results %>%
  select(Label, Receptor = Gene, logFC_Receptor = log2FC, pval_Receptor = adj.pvalue)
```

## Multi-comparison merge
Join network with ligands, then join with receptors matching botht he receptor name and the label.
Ensures evaluating ligand X and receptor Y within the same MSstats comparison

```{r}

all_cross_talk <- lr_with_evidence %>%
  #keeps only rows that exist in both
  inner_join(ligands_data, by = "Ligand", relationship = "many-to-many") %>%
  #match both receptor name and the label (condition of interest) - keeps only rows where the recieving recedptor was also detected, in that specific condition
  inner_join(receptors_data, by = c("Receptor", "Label")) %>%
  # Filter to keep only rows where BOTH proteins were detected and statistically significant
  filter(pval_Ligand < 0.05 & pval_Receptor < 0.05) %>%
  arrange(desc(curation_count))

#4839 length when inner joining to ligand
#2135 length when inner joining to receptor and label too.
#147 legnth when filtering to sig for both. 

```

##Extract co-upregulated pairs per comparison
```{r}
co_upregulated_all <- all_cross_talk %>%
  filter(logFC_Ligand > 0 & logFC_Receptor > 0) %>%
  group_by(Label) %>%
  arrange(desc(logFC_Ligand + logFC_Receptor)) %>%
  ungroup() %>%
  select(Label, Ligand, Receptor, logFC_Ligand, logFC_Receptor, paper_count, references)

co_upregulated_all
```

## check number of interactions for each comparison
```{r}
interaction_summary <- co_upregulated_all %>%
  count(Label, name = "Significant_Interactions")
print(interaction_summary)
#aplp2 cd14

```

```{r}
near_vs_ctrl <- all_cross_talk %>% 
  filter(Label == "PlaqueNear_vs_Control") %>%
  filter(pval_Ligand <0.05 & pval_Receptor <0.05) %>%
  arrange(desc(abs(logFC_Ligand + logFC_Receptor))) %>%
  select(-c(sources, references))

near_vs_ctrl

```

