---
title: "7. Putative cell-cell interactions"
format: html
---

==================================================================

# Cell-cell interactions for proteomics data

Author: Carlo Zanetti

This script aims to identify and statistically validate potential intercellular signalling pathways occurring between astrocytes in different plaques states compared to WT.

## Workflow overview

1.  **OmnipathR** package retrieves a database of **human** ligand-receptor interactions.
2.  Import DE results (derived from MSstats)
3.  Filter for cases were both the ligand and receptor are statistically significant (p \< 0.05) and upregulated (either logFC \> 0, or stricter 0.5, within the same experimental comparison).
4.  To ensure interactions aren't random occurrences, [1000 permutations]{.underline} are performed, shuffling the protein names to create a null distribution of random interactions. We can then compare the observed interaction counts to this distribution and derive an empirical p-value.
5.  Interactions are ranked and filtered by both the number of supporting research papers and combined magnitude of upregulation (`logFC_Ligand` + `logFC_Receptor`).

==================================================================

```{r}
suppressPackageStartupMessages({
library(OmnipathR)
library(dplyr)
library(tidyr)
library(data.table)
library(qs2)
library(ggplot2)
library(igraph)
library(ggraph)
library(patchwork)
})
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run5"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Retrieve intercellular communication database

```{r}
lr_network <- intercell_network(
  transmitter_param = list(categories = c('ligand')),
  receiver_param = list(categories = c('receptor')),
  resources = NULL,
  references_by_resource = TRUE)

lr_with_evidence <- lr_network %>%
    dplyr::select(
        Ligand = source_genesymbol, 
        Receptor = target_genesymbol, 
        sources, 
        references
    ) %>%
    distinct() %>%
    # Count how many databases confirm this interaction
    mutate(curation_count = nchar(as.character(sources)) - nchar(gsub(";", "", as.character(sources))) + 1) %>%
    # Count how many individual PMIDs (papers) support it
    mutate(paper_count = nchar(as.character(references)) - nchar(gsub(";", "", as.character(references))) + 1) %>%
    # Replace NAs with 0
    mutate(across(c(curation_count, paper_count), ~replace_na(.x, 0)))
```

## Read in MSstats results

```{r}
res_df <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))

clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))

protein_dictionary <- clean_spec_raw %>%
  dplyr::select(Protein = PG.ProteinGroups, 
         Gene = PG.Genes, 
         Description = PG.ProteinDescriptions) %>%
  distinct()

msstats_results <- res_df %>%
  # handle NAs
  filter(Label %in% c("PlaqueFar_vs_Control", "PlaqueNear_vs_Control")) %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  
  # Join with dictionary
  left_join(protein_dictionary, by = "Protein") %>%
  
  # Clean up Gene names and define significance
  mutate(
    # Fallback to Protein ID if Gene name is missing or empty
    #Change to upper case to match with human gene symbols
    Gene = toupper(ifelse(is.na(Gene) | Gene == "", Protein, Gene)),
  )
```

## Create ligand and receptor dataframes

```{r}
#Ligands data
ligands_data <- msstats_results %>%
  dplyr::select(Label, Ligand = Gene, logFC_Ligand = log2FC, pval_Ligand = adj.pvalue)

#Receptor data
receptors_data <- msstats_results %>%
  dplyr::select(Label, Receptor = Gene, logFC_Receptor = log2FC, pval_Receptor = adj.pvalue)
```

## Multi-comparison merge

Join network with ligands, then join with receptors matching both the receptor name and the label. Ensures evaluating ligand X and receptor Y within the same MSstats comparison.

```{r}

all_cross_talk <- lr_with_evidence %>%
  #keeps only rows that exist in both
  inner_join(ligands_data, by = "Ligand", relationship = "many-to-many") %>%
  #match both receptor name and the label (condition of interest) - keeps only rows where the recieving recedptor was also detected, in that specific condition
  inner_join(receptors_data, by = c("Receptor", "Label")) %>%
  # Filter to keep only rows where BOTH proteins were detected and statistically significant
  filter(pval_Ligand < 0.05 & pval_Receptor < 0.05) %>%
  arrange(desc(curation_count))

#4839 length when inner joining to ligand
#2135 length when inner joining to receptor and label too.
#147 length when filtering to sig for both. 

```

## Extract co-upregulated pairs per comparison (logFC \>0)

logFC filters set to 0 to be permissive.

```{r}
co_upregulated_all <- all_cross_talk %>%
  filter(logFC_Ligand > 0 & logFC_Receptor > 0) %>%
  group_by(Label) %>%
  arrange(desc(logFC_Ligand + logFC_Receptor)) %>%
  ungroup() %>%
  dplyr::select(Label, Ligand, Receptor, logFC_Ligand, logFC_Receptor, paper_count, references)

co_upregulated_all
```

## Comparison of number of interactions for each comparison

```{r}
interaction_summary <- co_upregulated_all %>%
  count(Label, name = "Observed_Interactions")
interaction_summary
```

## Permutation test (logFC \> 0)

```{r}
set.seed(42)
n_perms <- 1000
results_list <- vector("list", n_perms)

print(paste("Starting", n_perms, "permutations..."))

for (i in 1:n_perms) {
  
  if (i %% 100 == 0) {
    message("Completed ", i, " of ", n_perms, " permutations...")
  }
  
  # Shuffle Gene names to break biological links but keep data structure
  shuffled_msstats <- msstats_results %>%
    group_by(Label) %>%
    mutate(Gene = sample(Gene)) %>% 
    ungroup()
  
  # Rebuild fake ligand/receptor tables
  shuffled_ligands <- shuffled_msstats %>%
    dplyr::select(Label, Ligand = Gene, logFC_Ligand = log2FC, pval_Ligand = adj.pvalue)
  
  shuffled_receptors <- shuffled_msstats %>%
    dplyr::select(Label, Receptor = Gene, logFC_Receptor = log2FC, pval_Receptor = adj.pvalue)
  
  # Re-run the network merge and filtering with fake data
  shuffled_cross_talk <- lr_with_evidence %>%
    inner_join(shuffled_ligands, by = "Ligand", relationship = "many-to-many") %>%
    inner_join(shuffled_receptors, by = c("Receptor", "Label"), relationship = "many-to-many")%>%
    filter(pval_Ligand < 0.05, pval_Receptor < 0.05, 
           logFC_Ligand > 0, logFC_Receptor > 0)
  
  # Count hits
  shuffled_summary <- shuffled_cross_talk %>%
    count(Label, name = "Simulated_Interactions") %>%
    mutate(Permutation = i)
  
  results_list[[i]] <- shuffled_summary
}

# Combine simulated results
null_distribution_df <- bind_rows(results_list)

# Ensure cases where 0 interactions were found are explicitly tracked
all_labels <- unique(msstats_results$Label)
grid <- expand_grid(Label = all_labels, Permutation = 1:n_perms)
null_distribution_df <- grid %>%
  left_join(null_distribution_df, by = c("Label", "Permutation")) %>%
  mutate(Simulated_Interactions = replace_na(Simulated_Interactions, 0))

# Calculate Empirical P-values
empirical_results <- interaction_summary %>%
  left_join(null_distribution_df, by = "Label", relationship = "many-to-many") %>%
  group_by(Label, Observed_Interactions) %>%
  summarise(
    Expected_Mean = mean(Simulated_Interactions),
    Permutations_Greater_or_Equal = sum(Simulated_Interactions >= Observed_Interactions),
    Empirical_P_Value = sum(Simulated_Interactions >= Observed_Interactions) / n_perms,
    .groups = "drop"
  )
print(empirical_results)
```

## Visualize the Permutation Results

```{r}
perm_plot <- ggplot(null_distribution_df, aes(x = Simulated_Interactions)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black", alpha = 0.7) +
  geom_vline(data = interaction_summary, aes(xintercept = Observed_Interactions), 
             color = "red", linetype = "dashed", size = 1.2) +
  facet_wrap(~Label, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Permutation Test: Observed vs. Random Co-upregulated L-R Pairs",
    subtitle = "Red line = Actual observed pairs. Histogram = What random chance produced.",
    x = "Number of Significant Co-upregulated Pairs",
    y = "Frequency (out of 1000 permutations)"
  )

print(perm_plot)
```

## Extract co-upregulated pairs per comparison (logFC \>0.5)

Sorted by decreasing paper count

```{r}
co_upregulated_all <- all_cross_talk %>%
  filter(logFC_Ligand > 0.5 & logFC_Receptor > 0.5) %>%
  group_by(Label) %>%
  arrange(desc(paper_count)) %>%
  ungroup() %>%
  dplyr::select(Label, Ligand, Receptor, logFC_Ligand, logFC_Receptor, paper_count, references)

co_upregulated_all
```

Sorted by descending LogFC_Ligand + logFC_Receptor:

```{r}
co_upregulated_all <- all_cross_talk %>%
  filter(logFC_Ligand > 0.5 & logFC_Receptor > 0.5) %>%
  group_by(Label) %>%
  arrange(desc(logFC_Ligand + logFC_Receptor)) %>%
  ungroup() %>%
  dplyr::select(Label, Ligand, Receptor, logFC_Ligand, logFC_Receptor, paper_count, references)

co_upregulated_all
```

## Comparison of number of interactions for each comparison

```{r}
interaction_summary <- co_upregulated_all %>%
  count(Label, name = "Observed_Interactions")
interaction_summary
```

## Permutation test (logFC \> 0.5)

```{r}
set.seed(42)
n_perms <- 1000
results_list <- vector("list", n_perms)

print(paste("Starting", n_perms, "permutations..."))

for (i in 1:n_perms) {
  
  if (i %% 100 == 0) {
    message("Completed ", i, " of ", n_perms, " permutations...")
  }
  
  # Shuffle Gene names to break biological links but keep data structure
  shuffled_msstats <- msstats_results %>%
    group_by(Label) %>%
    mutate(Gene = sample(Gene)) %>% 
    ungroup()
  
  # Rebuild fake ligand/receptor tables
  shuffled_ligands <- shuffled_msstats %>%
    dplyr::select(Label, Ligand = Gene, logFC_Ligand = log2FC, pval_Ligand = adj.pvalue)
  
  shuffled_receptors <- shuffled_msstats %>%
    dplyr::select(Label, Receptor = Gene, logFC_Receptor = log2FC, pval_Receptor = adj.pvalue)
  
  # Re-run the network merge and filtering with fake data
  shuffled_cross_talk <- lr_with_evidence %>%
    inner_join(shuffled_ligands, by = "Ligand", relationship = "many-to-many") %>%
    inner_join(shuffled_receptors, by = c("Receptor", "Label"), relationship = "many-to-many")%>%
    filter(pval_Ligand < 0.05, pval_Receptor < 0.05, 
           logFC_Ligand > 0.5, logFC_Receptor > 0.5)
  
  # Count hits
  shuffled_summary <- shuffled_cross_talk %>%
    count(Label, name = "Simulated_Interactions") %>%
    mutate(Permutation = i)
  
  results_list[[i]] <- shuffled_summary
}

# Combine simulated results
null_distribution_df <- bind_rows(results_list)

# Ensure cases where 0 interactions were found are explicitly tracked
all_labels <- unique(msstats_results$Label)
grid <- expand_grid(Label = all_labels, Permutation = 1:n_perms)
null_distribution_df <- grid %>%
  left_join(null_distribution_df, by = c("Label", "Permutation")) %>%
  mutate(Simulated_Interactions = replace_na(Simulated_Interactions, 0))

# Calculate Empirical P-values
empirical_results <- interaction_summary %>%
  left_join(null_distribution_df, by = "Label", relationship = "many-to-many") %>%
  group_by(Label, Observed_Interactions) %>%
  summarise(
    Expected_Mean = mean(Simulated_Interactions),
    Permutations_Greater_or_Equal = sum(Simulated_Interactions >= Observed_Interactions),
    Empirical_P_Value = sum(Simulated_Interactions >= Observed_Interactions) / n_perms,
    .groups = "drop"
  )
print(empirical_results)
```

## Visualize the Permutation Results

```{r}
perm_plot <- ggplot(null_distribution_df, aes(x = Simulated_Interactions)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black", alpha = 0.7) +
  geom_vline(data = interaction_summary, aes(xintercept = Observed_Interactions), 
             color = "red", linetype = "dashed", size = 1.2) +
  facet_wrap(~Label, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Permutation Test: Observed vs. Random Co-upregulated L-R Pairs",
    subtitle = "Red line = Actual observed pairs. Histogram = What random chance produced.",
    x = "Number of Significant Co-upregulated Pairs",
    y = "Frequency (out of 1000 permutations)"
  )

print(perm_plot)
```


```{r}
# ==============================================================================
# Ligand-Receptor Interaction Heatmap (logFC > 0.5) - NEAR VS CONTROL ONLY
# Uses all_cross_talk object
# ==============================================================================


# ------------------------------------------------------------------------------
# 1. Prepare data — logFC > 0.5 co-upregulated pairs (PlaqueNear only)
# ------------------------------------------------------------------------------

lr_heatmap_data <- all_cross_talk %>%
  # Isolate only the PlaqueNear_vs_Control condition
  filter(Label == "PlaqueNear_vs_Control") %>%
  filter(logFC_Ligand > 0.5 & logFC_Receptor > 0.5) %>%
  mutate(
    combined_logFC = logFC_Ligand + logFC_Receptor,
    comparison = "Near vs Ctrl", 
    # Create a pair label for ordering
    pair = paste0(Ligand, " \u2192 ", Receptor)
  ) %>%
  dplyr::select(pair, Ligand, Receptor, comparison, 
                combined_logFC, logFC_Ligand, logFC_Receptor, paper_count)

# ------------------------------------------------------------------------------
# 2. Dot plot heatmap — pair x comparison
# ------------------------------------------------------------------------------

# Order pairs by combined logFC (descending)
pair_order <- lr_heatmap_data %>%
  group_by(pair) %>%
  summarise(max_fc = max(combined_logFC), .groups = "drop") %>%
  arrange(desc(max_fc)) %>%
  pull(pair)

lr_heatmap_data$pair <- factor(lr_heatmap_data$pair, levels = rev(pair_order))

p_dot <- ggplot(lr_heatmap_data, 
                aes(x = comparison, y = pair)) +
  geom_point(
    aes(size = paper_count + 1, fill = combined_logFC),
    shape = 21, colour = "black", stroke = 0.4
  ) +
  # Add logFC text inside/beside each dot
  geom_text(
    aes(label = sprintf("%.2f", combined_logFC)),
    size = 2.8, colour = "grey30", nudge_x = 0.3
  ) +
  scale_fill_gradient2(
    low = "lightyellow", mid = "#F0997B", high = "firebrick3",
    midpoint = median(lr_heatmap_data$combined_logFC),
    name = "Combined\nlog2FC"
  ) +
  scale_size_continuous(
    range = c(3, 12),
    name = "Paper\nevidence",
    breaks = c(1, 3, 7, 13),
    labels = c("0", "2", "6", "12")
  ) +
  labs(
    title = "Co-upregulated Ligand-Receptor Pairs",
    subtitle = "Near vs Control | Both log2FC > 0.5, adj.p < 0.05 | Dot size = supporting publications",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 10, colour = "grey40", margin = margin(b = 15)),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    plot.margin = margin(20, 20, 20, 10)
  )

print(p_dot)

ggsave(
  filename = file.path(results_dir, "LR_dotplot_Near_logFC05.pdf"),
  plot = p_dot,
  width = 8, height = 8, dpi = 300 # Slightly narrowed width for single column
)

# ------------------------------------------------------------------------------
# 3. Tile heatmap — ligand x receptor matrix
# ------------------------------------------------------------------------------

# Create a complete grid for the axes
all_ligands <- sort(unique(lr_heatmap_data$Ligand))
all_receptors <- sort(unique(lr_heatmap_data$Receptor))

tile_data <- lr_heatmap_data %>%
  dplyr::select(Ligand, Receptor, combined_logFC, paper_count) %>%
  # Dropped "comparison" from complete() since there is only one
  complete(Ligand = all_ligands, Receptor = all_receptors, 
           fill = list(combined_logFC = NA, paper_count = NA))

p_tile <- ggplot(tile_data, aes(x = Receptor, y = Ligand)) +
  geom_tile(
    aes(fill = combined_logFC),
    colour = "white", linewidth = 0.8
  ) +
  # Paper count as text overlay
  geom_text(
    aes(label = ifelse(!is.na(paper_count) & paper_count > 0, paper_count, "")),
    size = 3.5, fontface = "bold", colour = "white"
  ) +
  scale_fill_gradient(
    low = "#F5C4B3", high = "firebrick3",
    na.value = "grey95",
    name = "Combined\nlog2FC"
  ) +
  # Removed facet_wrap since we only have one condition to plot
  labs(
    title = "Ligand-Receptor Interaction Heatmap (Near vs Control)",
    subtitle = "log2FC > 0.5, adj.p < 0.05 | Numbers = supporting publications",
    x = "Receptor", y = "Ligand"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 10, colour = "grey40", margin = margin(b = 15)),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    axis.title = element_text(face = "bold"),
    panel.grid = element_blank(),
    legend.position = "right",
    plot.margin = margin(20, 20, 20, 10)
  )

print(p_tile)

ggsave(
  filename = file.path(results_dir, "LR_heatmap_tile_Near_logFC05.pdf"),
  plot = p_tile,
  width = 9, height = 7, dpi = 300 # Scaled down width for a single heatmap
)
```

