---
title: "07_cell_cell_interactions"
format: html
---

```{r}
library(OmnipathR)
library(dplyr)
library(tidyr)
library(data.table)
library(qs2)
library(ggplot2)
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run3"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Retrieve intercellular communication database

```{r}
lr_network <- intercell_network(
  transmitter_param = list(categories = c('ligand')),
  receiver_param = list(categories = c('receptor')),
  resources = NULL,
  references_by_resource = TRUE)

lr_with_evidence <- lr_network %>%
    dplyr::select(
        Ligand = source_genesymbol, 
        Receptor = target_genesymbol, 
        sources, 
        references
    ) %>%
    distinct() %>%
    # Count how many databases confirm this interaction
    mutate(curation_count = nchar(as.character(sources)) - nchar(gsub(";", "", as.character(sources))) + 1) %>%
    # Count how many individual PMIDs (papers) support it
    mutate(paper_count = nchar(as.character(references)) - nchar(gsub(";", "", as.character(references))) + 1) %>%
    # Replace NAs with 0
    mutate(across(c(curation_count, paper_count), ~replace_na(.x, 0)))
```

## Read in MSstats results

```{r}
res_df <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))

clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))

protein_dictionary <- clean_spec_raw %>%
  dplyr::select(Protein = PG.ProteinGroups, 
         Gene = PG.Genes, 
         Description = PG.ProteinDescriptions) %>%
  distinct()

msstats_results <- res_df %>%
  # handle NAs
  filter(Label %in% c("PlaqueFar_vs_Control", "PlaqueNear_vs_Control")) %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  
  # Join with dictionary
  left_join(protein_dictionary, by = "Protein") %>%
  
  # Clean up Gene names and define significance
  mutate(
    # Fallback to Protein ID if Gene name is missing or empty
    #Change to upper case to match with human gene symbols
    Gene = toupper(ifelse(is.na(Gene) | Gene == "", Protein, Gene)),
  )
```

## Create ligand and receptor dataframes

```{r}
#Ligands data
ligands_data <- msstats_results %>%
  dplyr::select(Label, Ligand = Gene, logFC_Ligand = log2FC, pval_Ligand = adj.pvalue)

#Receptor data
receptors_data <- msstats_results %>%
  dplyr::select(Label, Receptor = Gene, logFC_Receptor = log2FC, pval_Receptor = adj.pvalue)
```

## Multi-comparison merge

Join network with ligands, then join with receptors matching both the receptor name and the label. Ensures evaluating ligand X and receptor Y within the same MSstats comparison.

```{r}

all_cross_talk <- lr_with_evidence %>%
  #keeps only rows that exist in both
  inner_join(ligands_data, by = "Ligand", relationship = "many-to-many") %>%
  #match both receptor name and the label (condition of interest) - keeps only rows where the recieving recedptor was also detected, in that specific condition
  inner_join(receptors_data, by = c("Receptor", "Label")) %>%
  # Filter to keep only rows where BOTH proteins were detected and statistically significant
  filter(pval_Ligand < 0.05 & pval_Receptor < 0.05) %>%
  arrange(desc(curation_count))

#4839 length when inner joining to ligand
#2135 length when inner joining to receptor and label too.
#147 length when filtering to sig for both. 

```

## Extract co-upregulated pairs per comparison

```{r}
co_upregulated_all <- all_cross_talk %>%
  filter(logFC_Ligand > 0 & logFC_Receptor > 0) %>%
  group_by(Label) %>%
  arrange(desc(logFC_Ligand + logFC_Receptor)) %>%
  ungroup() %>%
  dplyr::select(Label, Ligand, Receptor, logFC_Ligand, logFC_Receptor, paper_count, references)

co_upregulated_all
```

## Comparison of number of interactions for each comparison

```{r}
interaction_summary <- co_upregulated_all %>%
  count(Label, name = "Observed_Interactions")
interaction_summary
```

## Permutation test

```{r}
set.seed(42)
n_perms <- 1000
results_list <- vector("list", n_perms)

print(paste("Starting", n_perms, "permutations..."))

for (i in 1:n_perms) {
  
  if (i %% 100 == 0) {
    message("Completed ", i, " of ", n_perms, " permutations...")
  }
  
  # Shuffle Gene names to break biological links but keep data structure
  shuffled_msstats <- msstats_results %>%
    group_by(Label) %>%
    mutate(Gene = sample(Gene)) %>% 
    ungroup()
  
  # Rebuild fake ligand/receptor tables
  shuffled_ligands <- shuffled_msstats %>%
    select(Label, Ligand = Gene, logFC_Ligand = log2FC, pval_Ligand = adj.pvalue)
  
  shuffled_receptors <- shuffled_msstats %>%
    select(Label, Receptor = Gene, logFC_Receptor = log2FC, pval_Receptor = adj.pvalue)
  
  # Re-run the network merge and filtering with fake data
  shuffled_cross_talk <- lr_with_evidence %>%
    inner_join(shuffled_ligands, by = "Ligand", relationship = "many-to-many") %>%
    inner_join(shuffled_receptors, by = c("Receptor", "Label"), relationship = "many-to-many")%>%
    filter(pval_Ligand < 0.05, pval_Receptor < 0.05, 
           logFC_Ligand > 0, logFC_Receptor > 0)
  
  # Count hits
  shuffled_summary <- shuffled_cross_talk %>%
    count(Label, name = "Simulated_Interactions") %>%
    mutate(Permutation = i)
  
  results_list[[i]] <- shuffled_summary
}

# Combine simulated results
null_distribution_df <- bind_rows(results_list)

# Ensure cases where 0 interactions were found are explicitly tracked
all_labels <- unique(msstats_results$Label)
grid <- expand_grid(Label = all_labels, Permutation = 1:n_perms)
null_distribution_df <- grid %>%
  left_join(null_distribution_df, by = c("Label", "Permutation")) %>%
  mutate(Simulated_Interactions = replace_na(Simulated_Interactions, 0))

# Calculate Empirical P-values
empirical_results <- interaction_summary %>%
  left_join(null_distribution_df, by = "Label", relationship = "many-to-many") %>%
  group_by(Label, Observed_Interactions) %>%
  summarise(
    Expected_Mean = mean(Simulated_Interactions),
    Permutations_Greater_or_Equal = sum(Simulated_Interactions >= Observed_Interactions),
    Empirical_P_Value = sum(Simulated_Interactions >= Observed_Interactions) / n_perms,
    .groups = "drop"
  )
print(empirical_results)
```

```{r}
# Visualize the Permutation Results 
perm_plot <- ggplot(null_distribution_df, aes(x = Simulated_Interactions)) +
  geom_histogram(binwidth = 1, fill = "lightgray", color = "black", alpha = 0.7) +
  geom_vline(data = interaction_summary, aes(xintercept = Observed_Interactions), 
             color = "red", linetype = "dashed", size = 1.2) +
  facet_wrap(~Label, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Permutation Test: Observed vs. Random Co-upregulated L-R Pairs",
    subtitle = "Red line = Actual observed pairs. Histogram = What random chance produced.",
    x = "Number of Significant Co-upregulated Pairs",
    y = "Frequency (out of 1000 permutations)"
  )

print(perm_plot)
```
