---
title: "08: PPI Network Analysis"
format: html
---

## Overview

This script identifies protein-protein interaction networks for upregulated
proteins in each MSstats comparison using STRINGdb. It identifies hub proteins
by degree centrality and exports summary tables for downstream interpretation.

**Workflow overview**

1. Load MSstats DE results and filter for upregulated proteins (FDR < 0.05, log2FC > 0.5)
2. Map gene names to STRING IDs via STRINGdb
3. Retrieve high-confidence interactions (STRING score >= 700)
4. Build and visualise full and top-N networks per comparison
5. Export hub protein summary tables across comparisons

## Libraries

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics')
```

```{r libraries, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(stringr)
library(STRINGdb)
library(igraph)
library(qs2)
})
```

## Directories and parameters

```{r directories}
base_dir    <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num     <- "run5"
results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
ppi_dir     <- file.path(results_dir, "PPI_networks")

dir.create(ppi_dir, recursive = TRUE, showWarnings = FALSE)

# Thresholds
FDR_CUTOFF   <- 0.05
LOGFC_CUTOFF <- 0.5
STRING_SCORE <- 700
```

## Load and prepare DE results

```{r load_data}
res_df <- fread(file.path(results_dir, "Full_GroupComparison_Results.csv"))

clean_spec_raw <- fread(file.path(base_dir, "data/processed/run1/clean_spec_raw.csv"))
protein_dictionary <- clean_spec_raw %>%
  dplyr::select(
    Protein     = PG.ProteinGroups,
    Gene        = PG.Genes,
    Description = PG.ProteinDescriptions
  ) %>%
  distinct()

# Map gene names and define significance categories
msstats_ready <- res_df %>%
  filter(!is.na(adj.pvalue), is.finite(log2FC)) %>%
  left_join(protein_dictionary, by = "Protein") %>%
  mutate(
    Gene = ifelse(is.na(Gene) | Gene == "", Protein, Gene),
    Status = case_when(
      adj.pvalue < FDR_CUTOFF & log2FC >  LOGFC_CUTOFF ~ "Upregulated",
      adj.pvalue < FDR_CUTOFF & log2FC < -LOGFC_CUTOFF ~ "Downregulated",
      TRUE ~ "Not Significant"
    )
  )
```

## Initialise STRINGdb

```{r init_string}
string_db <- STRINGdb$new(
  version         = "12.0",
  species         = 10090,
  score_threshold = STRING_SCORE,
  network_type    = "physical"
)
```

## Build networks per comparison

```{r build_networks}
comparisons <- c("PlaqueNear_vs_Control", "PlaqueFar_vs_Control", "PlaqueNear_vs_PlaqueFar")

network_results <- list()

for (comp in comparisons) {
  
  cat("\n============================\n")
  cat(comp, "\n")
  cat("============================\n")
  
  # --- Filter upregulated proteins ---
  sig_up <- msstats_ready %>%
    filter(Label == comp, Status == "Upregulated") %>%
    dplyr::select(Gene, log2FC, adj.pvalue) %>%
    distinct(Gene, .keep_all = TRUE) %>%
    as.data.frame()
  
  cat("Upregulated proteins:", nrow(sig_up), "\n")
  
  if (nrow(sig_up) < 5) {
    cat("Too few proteins, skipping.\n")
    next
  }
  
  # --- Map to STRING ---
  mapped <- string_db$map(sig_up, "Gene", removeUnmappedRows = TRUE)
  cat("Mapped to STRING:", nrow(mapped), "\n")
  
  if (nrow(mapped) < 5) {
    cat("Too few mapped, skipping.\n")
    next
  }
  
  # --- Get interactions ---
  interactions <- string_db$get_interactions(mapped$STRING_id)
  cat("Interactions:", nrow(interactions), "\n")
  
  # --- Build igraph object ---
  if (nrow(interactions) > 0) {
    
    g <- graph_from_data_frame(
      interactions %>% dplyr::select(from, to, combined_score),
      directed = FALSE,
      vertices = mapped %>% dplyr::select(STRING_id, Gene, log2FC)
    )
    g <- simplify(g)
    
    # Degree centrality
    V(g)$degree <- degree(g)
    
    cat("Nodes:", vcount(g), "| Edges:", ecount(g), "\n")
    
  } else {
    g <- NULL
    cat("No interactions found.\n")
  }
  
  network_results[[comp]] <- list(
    proteins     = mapped,
    interactions = interactions,
    graph        = g
  )
}
```

## Visualise full networks

```{r plot_full, fig.width=14, fig.height=14}
for (comp in names(network_results)) {
  
  g <- network_results[[comp]]$graph
  if (is.null(g) || ecount(g) == 0) next
  
  # Remove isolated nodes (degree 0) for cleaner plot
  g_plot <- delete_vertices(g, V(g)[degree(g) == 0])
  if (vcount(g_plot) < 3) next
  
  # Style
  V(g_plot)$label     <- V(g_plot)$Gene
  V(g_plot)$size      <- scales::rescale(degree(g_plot), to = c(5, 20))
  V(g_plot)$label.cex <- scales::rescale(degree(g_plot), to = c(0.6, 1.3))
  V(g_plot)$label.font <- 2
  V(g_plot)$color     <- "steelblue"
  V(g_plot)$frame.color <- "grey30"
  E(g_plot)$color     <- "grey75"
  E(g_plot)$width     <- 0.8
  
  # Highlight top 10 hubs
  top_hubs <- names(sort(degree(g_plot), decreasing = TRUE))[1:min(10, vcount(g_plot))]
  V(g_plot)$color[V(g_plot)$name %in% top_hubs] <- "firebrick3"
  
  pdf(file.path(ppi_dir, paste0("network_full_", comp, ".pdf")), width = 14, height = 14)
  set.seed(42)
  plot(g_plot,
       layout = layout_with_fr(g_plot),
       vertex.label.color = "black",
       main = paste0(comp, "\nUpregulated proteins | STRING physical >= ", STRING_SCORE,
                     "\nNodes: ", vcount(g_plot), " | Edges: ", ecount(g_plot),
                     " | Red = top 10 hubs"))
  dev.off()
  
  # Also render in notebook
  set.seed(42)
  plot(g_plot,
       layout = layout_with_fr(g_plot),
       vertex.label.color = "black",
       main = paste0(comp, "\nUpregulated proteins | STRING physical >= ", STRING_SCORE,
                     "\nNodes: ", vcount(g_plot), " | Edges: ", ecount(g_plot),
                     " | Red = top 10 hubs"))
}
```

## Visualise top-N subnetworks

Focused view of the most connected proteins only.

```{r plot_topN, fig.width=12, fig.height=12}
TOP_N <- 30

for (comp in names(network_results)) {
  
  g <- network_results[[comp]]$graph
  if (is.null(g) || ecount(g) == 0) next
  
  # Select top N by degree
  top_nodes <- names(sort(degree(g), decreasing = TRUE))[1:min(TOP_N, vcount(g))]
  g_top <- induced_subgraph(g, top_nodes)
  g_top <- delete_vertices(g_top, V(g_top)[degree(g_top) == 0])
  
  if (vcount(g_top) < 3) next
  
  V(g_top)$label      <- V(g_top)$Gene
  V(g_top)$size       <- scales::rescale(degree(g_top), to = c(8, 22))
  V(g_top)$label.cex  <- 1.0
  V(g_top)$label.font <- 2
  V(g_top)$frame.color <- "grey30"
  E(g_top)$color      <- "grey70"
  E(g_top)$width      <- 1.2
  
  # Colour by logFC
  fc <- V(g_top)$log2FC
  fc_pal <- colorRampPalette(c("gold", "orange", "firebrick3"))(100)
  fc_idx <- findInterval(fc, seq(min(fc), max(fc), length.out = 100))
  fc_idx[fc_idx == 0] <- 1
  V(g_top)$color <- fc_pal[fc_idx]
  
  pdf(file.path(ppi_dir, paste0("network_top", TOP_N, "_", comp, ".pdf")), 
      width = 12, height = 12)
  set.seed(42)
  plot(g_top,
       layout = layout_with_fr(g_top),
       vertex.label.color = "black",
       main = paste0(comp, " — Top ", vcount(g_top), " by degree",
                     "\nColoured by log2FC | Edges: ", ecount(g_top)))
  
  legend_vals <- seq(min(fc, na.rm = TRUE), max(fc, na.rm = TRUE), length.out = 5)
  legend_colors <- colorRampPalette(c("gold", "orange", "firebrick3"))(5)

  # 2. Add the legend
  legend("bottomright", 
         legend = round(legend_vals, 2), 
         fill = legend_colors, 
         title = "log2FC",
         bty = "n",       # No border box
         cex = 0.8)       # Text size
  dev.off()
  
  set.seed(42)
  plot(g_top,
       layout = layout_with_fr(g_top),
       vertex.label.color = "black",
       main = paste0(comp, " — Top ", vcount(g_top), " by degree",
                     "\nColoured by log2FC | Edges: ", ecount(g_top)))
  
  legend_vals <- seq(min(fc, na.rm = TRUE), max(fc, na.rm = TRUE), length.out = 5)
  legend_colors <- colorRampPalette(c("gold", "orange", "firebrick3"))(5)

  # 2. Add the legend
  legend("bottomright", 
         legend = round(legend_vals, 2), 
         fill = legend_colors, 
         title = "log2FC",
         bty = "n",       # No border box
         cex = 0.8)       # Text size
}
```

## Hub protein summary tables

```{r hub_tables}
hub_list <- list()

for (comp in names(network_results)) {
  
  g <- network_results[[comp]]$graph
  if (is.null(g)) next
  
  hub_df <- data.frame(
    Gene   = V(g)$Gene,
    log2FC = V(g)$log2FC,
    Degree = degree(g),
    Comparison = comp
  ) %>%
    filter(Degree > 0) %>%
    arrange(desc(Degree))
  
  hub_list[[comp]] <- hub_df
  
  cat("\n--- Top 20 hubs:", comp, "---\n")
  print(head(hub_df, 20))
  
  fwrite(hub_df, file.path(ppi_dir, paste0("hub_proteins_", comp, ".csv")))
}

# Combined table across comparisons
if (length(hub_list) > 0) {
  hub_combined <- bind_rows(hub_list)
  fwrite(hub_combined, file.path(ppi_dir, "hub_proteins_all_comparisons.csv"))
  
  # Proteins that are hubs in both comparisons
  if (length(hub_list) == 2) {
    shared_hubs <- inner_join(
      hub_list[[1]] %>% dplyr::select(Gene, Degree_Near = Degree, log2FC_Near = log2FC),
      hub_list[[2]] %>% dplyr::select(Gene, Degree_Far = Degree, log2FC_Far = log2FC),
      by = "Gene"
    ) %>%
      mutate(Total_Degree = Degree_Near + Degree_Far) %>%
      arrange(desc(Total_Degree))
    
    cat("\n--- Shared hubs (present in both comparisons) ---\n")
    print(head(shared_hubs, 20))
    
    fwrite(shared_hubs, file.path(ppi_dir, "hub_proteins_shared.csv"))
  }
}
```

## Network summary statistics

```{r summary_stats}
summary_table <- data.frame(
  Comparison           = character(),
  Upregulated_proteins = integer(),
  Mapped_to_STRING     = integer(),
  Nodes_with_edges     = integer(),
  Edges                = integer(),
  stringsAsFactors     = FALSE
)

for (comp in names(network_results)) {
  
  nr <- network_results[[comp]]
  g  <- nr$graph
  
  n_with_edges <- if (!is.null(g)) sum(degree(g) > 0) else 0
  n_edges      <- if (!is.null(g)) ecount(g) else 0
  
  summary_table <- rbind(summary_table, data.frame(
    Comparison           = comp,
    Upregulated_proteins = nrow(msstats_ready %>% filter(Label == comp, Status == "Upregulated") %>% distinct(Gene)),
    Mapped_to_STRING     = nrow(nr$proteins),
    Nodes_with_edges     = n_with_edges,
    Edges                = n_edges
  ))
}

cat("\n--- Network summary ---\n")
print(summary_table)
fwrite(summary_table, file.path(ppi_dir, "network_summary.csv"))
```

## Session info

```{r session_info}
sessionInfo()
```