---
title: "megena"
format: html
---

```{r}
library(qs2)
library(dplyr)
library(tidyr)
library(MEGENA)
library(doParallel)
library(org.Mm.eg.db)
library(tidyverse)
library(clusterProfiler)
library(ggplot2)
```

## Directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run1"

results_dir <- file.path(base_dir, "results", run_num, "MEGENA")
objects_dir <- file.path(base_dir, "data", "processed", run_num, "MEGENA")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
```

```{r}
# 1. Load Phase 1 processed data
processed_data <- qs_read(file.path(objects_dir, "processed_msstats_data.qs2"))
```

```{r}
prot_df <- as.data.frame(processed_data$ProteinLevelData)

wide_expr <- prot_df %>%
  select(originalRUN, Protein, LogIntensities) %>%
  pivot_wider(names_from = Protein, values_from = LogIntensities) %>%
  tibble::column_to_rownames("originalRUN")
```

# NA Filtering Strategy

```{r}

# Keep only proteins that are quantified in at least 80% of ALL your samples
threshold <- 0.5 * nrow(wide_expr)
keep_proteins <- colSums(!is.na(wide_expr)) >= threshold
filtered_expr <- wide_expr[, keep_proteins]

cat("Proteins retained for MEGENA:", ncol(filtered_expr), "\n")
```

## Impute

```{r}
expr_matrix <- apply(filtered_expr, 2, function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)
  return(x)
})
```

## MEGENA execution

```{r}
n_cores <- 32
registerDoParallel(cores = n_cores)

cat("Calculating Pearson correlation...\n")
cor_mat <- cor(expr_matrix, method = "pearson", use = "pairwise.complete.obs")

cat("Formatting into 3-column Edgelist for MEGENA...\n")
# CRITICAL FIX: Melt the matrix into a data.frame with Node1, Node2, and Weight
edgelist <- as.data.frame(as.table(cor_mat), stringsAsFactors = FALSE)
colnames(edgelist) <- c("Node1", "Node2", "Weight")

# Keep only unique pairs to remove self-loops and duplicates
edgelist <- edgelist[edgelist$Node1 < edgelist$Node2, ]

# Drop NAs
edgelist <- edgelist[!is.na(edgelist$Weight), ]

# PFN requires weights between 0 and 1, so we take the absolute correlation
edgelist$Weight <- abs(edgelist$Weight)

# Order by Weight descending (standard for PFN planarity testing)
edgelist <- edgelist[order(edgelist$Weight, decreasing = TRUE), ]
```

## MEGENA execution

```{r}
cat("Building Planar Filtered Network (PFN)...\n")
pfn_out <- calculate.PFN(edgelist, 
                         doPar = TRUE, 
                         num.cores = n_cores, 
                         keep.track = FALSE)


g_pfn <- graph_from_data_frame(pfn_out, directed = FALSE)

#Multiscale Clustering
megena_out <- do.MEGENA(g_pfn, 
                        mod.pval = 0.05, 
                        hub.pval = 0.05, 
                        remove.unsig = TRUE,
                        min.size = 10,   
                        max.size = 2000, 
                        doPar = TRUE, 
                        num.cores = n_cores, 
                        n.perm = 100,    
                        save.output = FALSE)
```

```{r}
qs_save(megena_out, file.path(objects_dir, "MEGENA_Full_Network.qs2"))
```

## Summary and saving

```{r}
cat("Summarizing modules...\n")
summary_out <- MEGENA.ModuleSummary(megena_out, 
                                    mod.pvalue = 0.05, 
                                    hub.pvalue = 0.05, 
                                    min.size = 10, 
                                    max.size = 2000, 
                                    annot.table = NULL)


module_table <- summary_out$modules

module_df <- enframe(module_table, name = "Module", value = "Protein") %>%
  unnest(Protein)



module_info <- summary_out$module.table

dir.create(file.path(results_dir, "MEGENA"), showWarnings = FALSE)
write.csv(module_df, file.path(objects_dir, "Protein_to_Module_Mapping.csv"), row.names = FALSE)
write.csv(module_info, file.path(objects_dir, "Module_Summary_Info.csv"), row.names = FALSE)

cat("Pipeline successfully finished.\n")
```

```{r}
print(length(megena_out$module))
```

```{r}
g_pfn <- graph_from_data_frame(pfn_out, directed = FALSE)
cat("Number of nodes in PFN:", vcount(g_pfn), "\n")
cat("Sample node names:", head(V(g_pfn)$name), "\n")

# Check if these names match your original matrix
cat("Do names match expr_matrix columns?", all(V(g_pfn)$name %in% colnames(expr_matrix)), "\n")
```

```{r}
raw_modules <- megena_out$module
all_res <- list()

# MEGENA stores modules by 'Scale'. We iterate through scales to find the modules.
for(scale_idx in 1:length(raw_modules)) {
    scale_data <- raw_modules[[scale_idx]]
    for(mod_id in names(scale_data)) {
        all_res[[mod_id]] <- data.frame(
            Protein = scale_data[[mod_id]],
            Module = mod_id,
            Scale = scale_idx
        )
    }
}

final_roster <- do.call(rbind, all_res)

cat("Successfully recovered", nrow(final_roster), "protein-module assignments.\n")
head(final_roster)

protein_to_module_mapping <- final_roster
write.csv(protein_to_module_mapping, file.path(objects_dir, "MEGENA", "Protein_to_Module_Mapping.csv"), row.names = FALSE)
```

```{r}
protein_map <- read.csv(file.path(objects_dir, "Protein_to_Module_Mapping.csv"))
protein_map$Gene_Symbol <- mapIds(org.Mm.eg.db, keys = protein_map$Protein, 
                                  column = "SYMBOL", keytype = "UNIPROT", multiVals = "first")
protein_map

# 2. Map the Module Summary (Hubs)
module_summary <- read.csv(file.path(objects_dir, "Module_Summary_Info.csv"))

# Function to convert hub string "ID(degree)" to "Symbol(degree)"
translate_hubs <- function(hub_string) {
  if (is.na(hub_string) || hub_string == "") return(NA)
  uids <- str_extract_all(hub_string, "[A-Z0-9]+")[[1]]
  uids <- uids[!grepl("^\\d+$", uids)] # Keep only alpha-numeric IDs
  syms <- mapIds(org.Mm.eg.db, keys = uids, column = "SYMBOL", 
                 keytype = "UNIPROT", multiVals = "first")
  new_string <- hub_string
  for (i in seq_along(uids)) {
    if (!is.na(syms[i])) new_string <- str_replace(new_string, uids[i], syms[i])
  }
  return(new_string)
}

module_summary$module.hub.genes <- sapply(module_summary$module.hub, translate_hubs)
module_summary
# Save the enriched files
write.csv(protein_map, file.path(objects_dir, "Protein_to_Module_Mapping_Final.csv"), row.names = FALSE)
write.csv(module_summary, file.path(objects_dir, "Module_Summary_Final.csv"), row.names = FALSE)
```

## MEGENA enrichment of modules

```{r}
clean_map <- subset(protein_map, !is.na(Gene_Symbol))
clean_map
module_go_enrichment <- compareCluster(Gene_Symbol ~ Module, 
                                       data = clean_map, 
                                       fun = "enrichGO",
                                       OrgDb = org.Mm.eg.db,
                                       keyType = "SYMBOL",
                                       ont = "BP",           # Biological Process
                                       pAdjustMethod = "BH", # Benjamini-Hochberg correction
                                       pvalueCutoff = 0.05,
                                       qvalueCutoff = 0.05)


enrichment_plot <- dotplot(module_go_enrichment, showCategory = 3) + 
                   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 4. Save the plot and the data (crucial for HPC workflows)
ggsave(file.path(results_dir, "MEGENA_Module_GO_Enrichment.pdf"), plot = enrichment_plot, width = 32, height = 20)
```

## try collapsing redunadant terms based on semantic similarity

```{r}
# Simplify the GO terms by removing redundant pathways (cutoff = 0.7 is standard)
simplified_go <- simplify(module_go_enrichment, cutoff = 0.7, by = "p.adjust", select_fun = min)

# Plot the simplified version
dotplot(simplified_go, showCategory = 3) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
leaf_modules <- setdiff(unique(module_summary$module.id), unique(module_summary$module.parent))
leaf_map <- subset(clean_map, Module %in% leaf_modules)
leaf_map

module_go_enrichment <- compareCluster(Gene_Symbol ~ Module, 
                                       data = leaf_map, 
                                       fun = "enrichGO",
                                       OrgDb = org.Mm.eg.db,
                                       keyType = "SYMBOL",
                                       ont = "BP",           # Biological Process
                                       pAdjustMethod = "BH", # Benjamini-Hochberg correction
                                       pvalueCutoff = 0.05,
                                       qvalueCutoff = 0.05)


enrichment_plot <- dotplot(module_go_enrichment, showCategory = 3) + 
                   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 4. Save the plot and the data (crucial for HPC workflows)
ggsave(file.path(results_dir, "MEGENA_Module_GO_Enrichment_leaf.pdf"), plot = enrichment_plot, width = 32, height = 20)
```

## Calculate module eigengenes

```{r}
eigengenes <- data.frame(row.names = rownames(expr_matrix))
expr_matrix
# Get list of all MEGENA modules
unique_modules <- unique(protein_map$Module)
protein_map$Gene_Symbol
# Loop through each module and calculate PC1
for(mod in unique_modules) {
  mod_proteins <- subset(protein_map, Module == mod)$Protein
  # Get the gene symbols for this specific module
  
  # Ensure the genes actually exist in your expression matrix
  valid_genes <- intersect(mod_proteins, colnames(expr_matrix))
  
  # Only run PCA if the module has enough valid genes
  if(length(valid_genes) > 3) { 
    mod_expr <- expr_matrix[, valid_genes]
    
    # Run PCA (scaling is crucial here so highly expressed genes don't dominate)
    pca <- prcomp(mod_expr, scale. = TRUE)
    
    # Extract PC1 (The Eigengene) and store it
    eigengenes[[mod]] <- pca$x[, 1]
  }
}

# View the resulting matrix (Rows = Samples, Cols = Module Eigengenes)
head(eigengenes)
```

```{r}
library(WGCNA)
library(pheatmap)


metadata <- read.csv(file.path(base_dir, "data", "processed", run_num, "msstats_annotation.csv"))
metadata
```

## Prep data for modelling

Want traits and eigengenes in one df.

Set control as baseline reference level.

```{r}
#bind eigengenes and metadata together
model_data <- cbind(eigengenes, metadata)

# 2. Force Condition to be a factor and set "control" as the baseline reference
model_data$Condition <- factor(model_data$Condition, 
                               levels = c("control", "plaque_far", "plaque_near"))

# 3. Ensure BioReplicate is a factor (for the random effect)
model_data$BioReplicate <- as.factor(model_data$BioReplicate)
```

## Run mixed-effects loop

```{r}
library(lme4)
library(lmerTest)
modules <- colnames(eigengenes)
t_matrix <- matrix(NA, nrow = length(modules), ncol = 2, 
                   dimnames = list(modules, c("Plaque_Far", "Plaque_Near")))
p_matrix <- t_matrix 

# Loop through each module
for(mod in modules) {
  
  # The formula: Module Expression ~ Condition + (Random Effect of Mouse)
  formula_str <- paste0(mod, " ~ Condition + (1 | BioReplicate)")
  
  # Fit the model
  fit <- lmer(as.formula(formula_str), data = model_data)
  
  # Extract the statistical summary
  coefs <- summary(fit)$coefficients
  
  # Store the t-values (Estimate / Std. Error)
  t_matrix[mod, "Plaque_Far"]  <- coefs["Conditionplaque_far", "t value"]
  t_matrix[mod, "Plaque_Near"] <- coefs["Conditionplaque_near", "t value"]
  
  # Store the p-values
  p_matrix[mod, "Plaque_Far"]  <- coefs["Conditionplaque_far", "Pr(>|t|)"]
  p_matrix[mod, "Plaque_Near"] <- coefs["Conditionplaque_near", "Pr(>|t|)"]
}
```

## Plot mixed effects model heatmap
Random effect is hijacking fixed effect. As control samples condition only ever comes from control mice - massive change atributes to random effect rather than plaque_near condition. 
```{r}
# 1. Create the text overlay (showing the t-value and the p-value)
text_matrix <- paste0("t=", signif(t_matrix, 2), "\n(p=", signif(p_matrix, 2), ")")
dim(text_matrix) <- dim(t_matrix)

# 2. Find the max absolute t-value to center our color scale symmetrically at zero
max_t <- max(abs(t_matrix), na.rm = TRUE)
breaks_list <- seq(-max_t, max_t, length.out = 51)

# 3. Plot the heatmap
lmm_plot <- pheatmap(t_matrix, 
                     display_numbers = text_matrix,
                     color = colorRampPalette(c("blue", "white", "red"))(50),
                     breaks = breaks_list,  # Ensures 0 is perfectly white
                     cluster_cols = FALSE,  # Keep Far -> Near logical order
                     cluster_rows = TRUE,  
                     main = "LMM T-Values: Module Expression vs Control",
                     fontsize_number = 8)

lmm_plot
# Save the plot
ggsave(file.path(results_dir, "LMM_Module_Trait_Heatmap.pdf"), plot = lmm_plot, width = 6, height = 12)
```


```{r}
# 1. Subset the data to ONLY include the disease mice (drop the controls)
disease_data <- subset(model_data, Condition != "control")

# 2. Re-level the Condition factor so plaque_far is the baseline/intercept
disease_data$Condition <- factor(disease_data$Condition, 
                                 levels = c("plaque_far", "plaque_near"))

# 3. Initialize matrices for the single comparison (Near vs Far)
modules <- colnames(eigengenes)
t_matrix_spatial <- matrix(NA, nrow = length(modules), ncol = 1, 
                           dimnames = list(modules, "Near_vs_Far"))
p_matrix_spatial <- t_matrix_spatial

# 4. Run the Mixed-Effects Loop on the purely spatial data
for(mod in modules) {
  # Formula: Expression ~ Condition + (1 | BioReplicate)
  fit <- lmer(as.formula(paste0(mod, " ~ Condition + (1 | BioReplicate)")), 
              data = disease_data)
  
  coefs <- summary(fit)$coefficients
  
  # Extract t-value and p-value for the plaque_near shift
  t_matrix_spatial[mod, "Near_vs_Far"] <- coefs["Conditionplaque_near", "t value"]
  p_matrix_spatial[mod, "Near_vs_Far"] <- coefs["Conditionplaque_near", "Pr(>|t|)"]
}
```


```{r}
# 1. Create the text overlay (showing the t-value and the p-value)
text_matrix_spatial <- paste0("t=", signif(t_matrix_spatial, 2), "\n(p=", signif(p_matrix_spatial, 2), ")")
dim(text_matrix_spatial) <- dim(t_matrix_spatial)

# 2. Find the max absolute t-value to center our color scale symmetrically at zero
max_t_spatial <- max(abs(t_matrix_spatial), na.rm = TRUE)
breaks_list_spatial <- seq(-max_t_spatial, max_t_spatial, length.out = 51)

# 3. Open a PDF device to save the plot directly to your results folder
pdf(file.path(results_dir, "Spatial_LMM_Heatmap_Near_vs_Far.pdf"), width = 6, height = 12)

# 4. Generate the single-column heatmap
pheatmap(t_matrix_spatial, 
         display_numbers = text_matrix_spatial,
         color = colorRampPalette(c("blue", "white", "red"))(50),
         breaks = breaks_list_spatial,
         cluster_cols = FALSE,  # Only one column, no need to cluster
         cluster_rows = TRUE,   # Groups similarly behaving modules together
         main = "LMM T-Values: Plaque Near vs Plaque Far",
         fontsize_number = 8,
         cellwidth = 100)       # Forces the column to be wide enough to read the text

# 5. Close the device and save
dev.off()
```

```{r}
# 1. Identify all modules that act as a parent
parent_modules <- unique(module_info$module.parent) # Use your loaded mapping df
parent_modules
#The "Leaf" modules (modules that never appear in the parent column)
leaf_modules <- setdiff(rownames(t_matrix_spatial), parent_modules)

# 3. Subset your t-matrix and p-matrix to ONLY include these leaves
t_matrix_leaves <- t_matrix_spatial[leaf_modules, , drop = FALSE]
p_matrix_leaves <- p_matrix_spatial[leaf_modules, , drop = FALSE]

# 4. Re-create the text matrix for the leaves
text_matrix_leaves <- paste0("t=", signif(t_matrix_leaves, 2), "\n(p=", signif(p_matrix_leaves, 2), ")")
dim(text_matrix_leaves) <- dim(t_matrix_leaves)

# 5. Re-plot the much cleaner heatmap
pheatmap(t_matrix_leaves, 
         display_numbers = text_matrix_leaves,
         color = colorRampPalette(c("blue", "white", "red"))(50),
         breaks = breaks_list_spatial, # Keep the same color scale as before
         cluster_cols = FALSE,  
         cluster_rows = TRUE,   
         main = "Leaf Modules: Plaque Near vs Far",
         fontsize_number = 8,
         cellwidth = 100)
```

```{r}
target_matrix <- t_matrix_spatial # change to t_matrix_leaves if you prefer

# 1. Find the Highest Upregulated Module (Deepest Red)
top_up_module <- rownames(target_matrix)[which.max(target_matrix[, "Near_vs_Far"])]
top_up_t <- max(target_matrix[, "Near_vs_Far"])

# 2. Find the Most Downregulated Module (Deepest Blue)
top_down_module <- rownames(target_matrix)[which.min(target_matrix[, "Near_vs_Far"])]
top_down_t <- min(target_matrix[, "Near_vs_Far"])

# 3. Print the results to your console so you can see them immediately
cat("Most Upregulated at Plaque:", top_up_module, "(t =", top_up_t, ")\n")
cat("Most Downregulated at Plaque:", top_down_module, "(t =", top_down_t, ")\n")
```

```{r}

# 1. Subset your clean gene map for ONLY these two extreme modules
extreme_modules <- c(top_up_module, top_down_module)
extreme_genes <- subset(protein_map, Module %in% extreme_modules & !is.na(Gene_Symbol))
extreme_genes
# 2. Run the GO Enrichment
extreme_go <- compareCluster(Gene_Symbol ~ Module, 
                             data = extreme_genes, 
                             fun = "enrichGO",
                             OrgDb = org.Mm.eg.db,
                             keyType = "SYMBOL",
                             ont = "BP", # Biological Process
                             pAdjustMethod = "BH",
                             pvalueCutoff = 0.05)

# 3. Plot and save to your HPC
extreme_plot <- dotplot(extreme_go, showCategory = 10) + 
                theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                ggtitle("Biological Functions of Extreme Plaque Modules")

ggsave(file.path(results_dir, "Top_and_Bottom_Module_GO.pdf"), plot = extreme_plot, width = 10, height = 8)
```


```{r}
unique_modules <- unique(clean_map$Module)

# 2. Define the background "universe" of genes
# This is statistically crucial: it tells the algorithm to only compare against 
# the proteins actually detected in your mass spec, not the entire mouse genome.
background_genes <- unique(clean_map$Gene_Symbol)

# 3. Open the PDF device 
# Everything printed after this line goes into a new page of this PDF
pdf_path <- file.path(results_dir, "MEGENA_All_Modules_Enrichment_Report.pdf")
pdf(file = pdf_path, width = 10, height = 8)

# 4. Loop through every module
for (mod in unique_modules) {
  
  # Extract the genes for this specific module
  mod_genes <- subset(clean_map, Module == mod)$Gene_Symbol
  
  # Run the GO Enrichment
  ego <- enrichGO(gene          = mod_genes,
                  universe      = background_genes,
                  OrgDb         = org.Mm.eg.db,
                  keyType       = "SYMBOL",
                  ont           = "BP",           # Biological Process
                  pAdjustMethod = "BH",
                  pvalueCutoff  = 0.05)
  
  # 5. The Safety Check: Does the enrichment have actual results?
  if (!is.null(ego) && nrow(as.data.frame(ego)) > 0) {
    
    # ---------------------------------------------------------
    # NEW STEP: Aggregate the redundant terms
    # cutoff = 0.7 means if two terms are 70% similar, keep the most significant one
    # ---------------------------------------------------------
    ego_simplified <- simplify(ego, 
                               cutoff = 0.7, 
                               by = "p.adjust", 
                               select_fun = min)
    
    # Second safety check just in case simplify() removed everything (rare)
    if (nrow(as.data.frame(ego_simplified)) > 0) {
      
      # Plot the CLEANED data
      p <- dotplot(ego_simplified, showCategory = 15) + 
           ggtitle(paste("GO Enrichment for Module:", mod, 
                         "| Size:", length(mod_genes), "Proteins")) +
           theme(plot.title = element_text(face = "bold", size = 14))
      
      print(p)
      
    } else {
      # Fallback if simplify() results in an empty dataframe
      p_empty <- ggplot() + 
                 annotate("text", x = 0.5, y = 0.5, size = 6,
                          label = paste("Module:", mod, "\nNo significant distinct GO terms.")) + 
                 theme_void()
      print(p_empty)
    }
    
  } else {
    
    # If no significant terms, create a blank placeholder plot
    p_empty <- ggplot() + 
               annotate("text", x = 0.5, y = 0.5, size = 6,
                        label = paste("Module:", mod, "\nNo significant GO terms found.")) + 
               theme_void()
    
    print(p_empty)
  }
}

# 6. Close the PDF device to finalize and save the file
dev.off()

cat("Enrichment report successfully saved to:", pdf_path, "\n")
```

```{r}
target_module <- "c1_55"
genes_in_module <- final_roster$Protein[final_roster$Module == target_module]

cat("There are", length(genes_in_module), "genes in", target_module, "\n")
print(head(genes_in_module, 20))
```

```{r}
summary_info <- MEGENA.ModuleSummary(megena_out, 
                                     mod.pvalue = 0.05, 
                                     hub.pvalue = 0.05, 
                                     min.size = 10)

# View the table of modules with their designated Hubs
module_table <- summary_info$module.table
print(head(module_table[, c("id", "module.size", "module.hub")]))
```

```{r}
spatial_sorted <- t_matrix_spatial[order(t_matrix_spatial[,1], decreasing = TRUE), , drop = FALSE]

# 2. Pick the Top 3 Upregulated and Top 3 Downregulated
top_3_up   <- rownames(head(spatial_sorted, 3))
top_3_down <- rownames(tail(spatial_sorted, 3))

# 3. View the Gene Symbols for the most Upregulated module
# (Let's look at the #1 upregulated module)
top_mod_id <- top_3_up[1]
top_genes  <- subset(clean_map, Module == top_mod_id)$Gene_Symbol

cat("### Genes in the Top Upregulated Module (", top_mod_id, ") ###\n")
print(sort(top_genes))

# 4. View the Gene Symbols for the most Downregulated module
down_mod_id <- top_3_down[3]
down_genes  <- subset(clean_map, Module == down_mod_id)$Gene_Symbol

cat("\n### Genes in the Top Downregulated Module (", down_mod_id, ") ###\n")
print(sort(down_genes))
```

