---
title: "Protein Normalization & PCA"
---

## Libraries

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics')
library(data.table)
library(MSstats)
library(qs2)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(pheatmap)
```

## Set directories

```{r}
base_dir <- "/nemo/lab/destrooperb/home/shared/zanettc/millie_proteomics"
run_num <- "run1"

results_dir <- file.path(base_dir, "results", run_num)
objects_dir <- file.path(base_dir, "data", "processed", run_num)
raw_dir <- file.path(base_dir, "data", "raw")

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(objects_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
```

## Read in previous files

```{r}
clean_spec_raw <- fread(file.path(objects_dir, "clean_spec_raw.csv"))
# Separate annotations from quantitative data
anno_cols <- c("PG.ProteinGroups", "PG.Genes", "PG.ProteinDescriptions")
quant_only <- clean_spec_raw[, !..anno_cols]
annotations <- clean_spec_raw[, ..anno_cols]

qc_data <- fread(file.path(objects_dir, "clean_metadata.csv"))
```

## Log transform on filtered dataset

Log2 Transform the filtered data (adding +1 to handle zeros)

Have to log transform because median normalisation is additive - subtracts median differnece from whole sample to shift up or down.

If you add or subtract a constant value from raw data -\> distort biological ratios.

By transforming to log2 first -\> multiplicative relationships become additive ones because of log laws.

Also tames dynamic range. Ensures hyper-abundant proteins don't dominate the maths.

Log transforming also makes the right-skewed distribution into a more normalish distribution, making the median a more accurate reflection of the whole sample.

```{r}
clean_quant_log <- log2(quant_only + 1)
#converts back to NA for accurate medians
clean_quant_log[clean_quant_log == 0] <- NA
```

## Median normalisation

```{r}
sample_meds <- apply(clean_quant_log, 2, median, na.rm = TRUE)

#Calculate the global median across the entire filtered experiment
global_med <- median(sample_meds)

#Calculate the shift factor for each sample
norm_factors <- global_med - sample_meds

# Apply the shift to normalize the data
quant_norm <- sweep(clean_quant_log, 2, norm_factors, "+")
```

```{r}
boxplot(quant_norm, outline = FALSE, 
        main = "Post-Normalization Distributions (Cleaned Data)", 
        las = 2, cex.axis = 0.4, col = "skyblue")
abline(h = global_med, col = "red", lwd = 2)
```

```{r}
# 1. Convert the normalized wide matrix to a data.frame and make it 'long'
abund_norm_long <- as.data.frame(quant_norm) %>%
  mutate(Protein = row_number()) %>%
  pivot_longer(-Protein, names_to = "Run", values_to = "Log2_Intensity") %>%
  drop_na()

# 2. Merge with metadata for the batch colors
# We use qc_data as the source for Batch and other metadata
abund_norm_plot_data <- abund_norm_long %>%
  left_join(qc_data, by = "Run")

# 3. Create the Plot
p_abund_norm <- ggplot(abund_norm_plot_data, aes(x = reorder(Run, Batch), y = Log2_Intensity, fill = as.factor(Batch))) +
  # All samples here are 'Pass' because we filtered the matrix earlier
  geom_boxplot(outlier.shape = NA, lwd = 0.4, color = "black") +
  theme_minimal() +
  labs(title = "Log2 Abundance per Run (Post-Normalization)",
       subtitle = "All runs are now aligned by their medians to remove batch/loading bias",
       x = "Clean Runs (Ordered by Batch)", 
       y = "Normalized Log2 Intensity", 
       fill = "Batch") +
  theme(axis.text.x = element_blank(),
        panel.grid.major.x = element_blank()) # Cleaner look with many samples

print(p_abund_norm)
```

## PCA on normalised data

```{r}
## PCA on Normalized Data
cat("Preparing Normalized Data for PCA...\n")

# 1. Transpose the normalized matrix so Runs = Rows, Proteins = Columns
# PCA expects samples as rows
pca_norm_mat <- t(quant_norm)

# 2. Handle remaining NAs (Imputation)
# Since we already log-transformed and normalized, we'll use a 
# simple downshifted imputation for the remaining missing values
# so prcomp doesn't fail.
set.seed(42)
mu <- mean(pca_norm_mat, na.rm = TRUE)
sigma <- sd(pca_norm_mat, na.rm = TRUE)
pca_norm_mat[is.na(pca_norm_mat)] <- rnorm(sum(is.na(pca_norm_mat)), 
                                           mean = mu - 1.8 * sigma, 
                                           sd = 0.3 * sigma)

# 3. Filter for Zero Variance
# Some proteins might become flat lines after normalization/imputation
valid_cols <- apply(pca_norm_mat, 2, var) > 0
pca_norm_mat <- pca_norm_mat[, valid_cols]

# 4. Run prcomp
pca_norm_res <- prcomp(pca_norm_mat, center = TRUE, scale. = TRUE)

# 5. Extract Coordinates and Merge Metadata
pca_norm_df <- as.data.frame(pca_norm_res$x)
pca_norm_df$Run <- rownames(pca_norm_df)
pca_norm_df <- merge(pca_norm_df, qc_data, by = "Run")

# Calculate Variance Explained
var_exp <- round(100 * pca_norm_res$sdev^2 / sum(pca_norm_res$sdev^2), 1)

# ---------------------------------------------------------
# Plot A: PCA by Condition (Biological Check)
# ---------------------------------------------------------
p_pca_cond <- ggplot(pca_norm_df, aes(x = PC1, y = PC2, color = Condition)) +
  geom_point(size = 4, alpha = 0.8) +
  theme_minimal() +
  scale_color_manual(values = c("control" = "#1b9e77", "plaque_near" = "#d95f02", "plaque_far" = "#7570b3")) +
  labs(title = "PCA of Normalized Data (Color by Condition)",
       subtitle = "If normalization worked, biological groups should start to cluster",
       x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"))

# ---------------------------------------------------------
# Plot B: PCA by Batch (Technical Check for Batch 6)
# ---------------------------------------------------------
p_pca_batch <- ggplot(pca_norm_df, aes(x = PC1, y = PC2, color = as.factor(Batch))) +
  geom_point(size = 4, alpha = 0.8) +
  theme_minimal() +
  labs(title = "PCA of Normalized Data (Color by Batch)",
       subtitle = "Checking if Batch 6 still drives the separation",
       x = paste0("PC1 (", var_exp[1], "%)"),
       y = paste0("PC2 (", var_exp[2], "%)"),
       color = "Batch")

print(p_pca_cond)
print(p_pca_batch)
```
